---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ggheat

<!-- badges: start -->
[![R-CMD-check](https://github.com/Yunuuuu/ggheat/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/Yunuuuu/ggheat/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

This package is a `ggplot2` extension for heatmap. It utilizes the grammar of
graphics to construct the heatmap and heatmap annotations.

## Installation

You can install the development version of ggheat from [GitHub](https://github.com/) with:

```{r, eval=FALSE}
# install.packages("devtools")
devtools::install_github("Yunuuuu/ggheat")
```


```{r setup}
library(ggplot2)
library(ggheat)
```

Let's begin by creating some example data

```{r setup_dat}
set.seed(123)
small_mat <- matrix(rnorm(81), nrow = 9)
rownames(small_mat) <- paste0("row", seq_len(nrow(small_mat)))
colnames(small_mat) <- paste0("column", seq_len(ncol(small_mat)))
```

## `ggheat`
The main function is `ggheat`, it's just like the ggplot function, which set the
default data and mapping. Using `ggheat`, it is effortless to create a simple
Heatmap. 

```{r}
ggheat(small_mat)
```

The data can be a numeric or character vector, a data frame, and any other data
which can be converted into a matrix. Simple vector will be converted into a one
column matrix. They will all converted into the long formated data frame when
drawing. The default mapping will use `aes(.data$.x, .data$.y)`, you can use
`mapping` argument to control it. The data contains following columns:  

   - `.row_panel` or `.column_panl` instead of a column `.panel` since
   annotation can only have one facet axiss: the row panel  
   
   - `.column_panel`: the column panel
   
   - `.row_names` and `.column_names`: the row and column names of the original
     matrix (only applicable when names exist).
   
   - `.row_index` and `.column_index`: the row and column index of the original
     matrix.
   
   - `.x` and `.y`: the `x` and `y` coordinates
   
   - `value`: the actual matrix value.

You can then add all other ggplot2 elements like `geoms`, `scales` and `facets`.

```{r}
ggheat(letters)
```

```{r}
ggheat(1:10)
```

```{r}
ggheat(small_mat) + scale_fill_viridis_c()
```

You can turn off the default filling by setting `filling = FALSE`
```{r}
ggheat(small_mat, filling = FALSE) +
  scale_fill_viridis_c() +
  geom_text(aes(label = sprintf("%d * %d", .row_index, .column_index)))
```

The position scales and facets are partial support.

For position scales, you cannot set `limits`, `breaks`, `labels`, and they are
just ignored. `limits` cannot be touched, and the internal will always reset it
as the default. This is required by heatmap annotations. But you can set
`labels` or `breaks` in `ggheat()` function. See `xlabels/ylabels` and
`xlabels_nudge/ylabels_nudge`.  All of these arguments should be provided in the
original order of the raw matrix. Even if we'll use heatmap annotation to
reorder the heatmap rows/columns. 
```{r}
ggheat(small_mat) + scale_x_continuous(limits = c(0, 0))
```

```{r}
ggheat(small_mat) + scale_x_continuous(labels = "a")
```

```{r}
ggheat(small_mat, xlabels = rep_len("AA", ncol(small_mat)))
```

When working with facets, manual configuration of panels using the `facet_*`
function is not possible. However, the `facet_*` function can still be used to
control the `facet` looks. 

To move forward with the subsequent procedures, it is necessary to introduce the
heatmap annotation first. Heatmap annotations are essential components that
contribute to the usability of a heatmap. 


## `htanno`

`ggheat` offers two primary annotations: `htanno` and `gganno`. 

`htanno` is a special annotation that can interact with the main heatmap. It
particularly allows for controlling the order of the main heatmap or dividing
the heatmap into sections based on rows/columns. 

Heatmap annotations can be positioned at the `top`, `left`, `bottom`, or `right`
of the heatmap. This is referred to as the active component in `ggheat`. Each
annotation function provides a position argument to control the placement of the
annotation. 

Currently, there are only two classes available for `htanno`: `htanno_dendro`
and `htanno_group`. The `htanno_dendro` class allows the addition of a
dendrogram near the heatmap. This is primarily useful when working with heatmap
plots. The primary purpose of a dendrogram is to facilitate clustering and
it can also reorder the heatmap. 

```{r htanno_dendro}
ggheat(small_mat) + htanno_dendro(position = "top")
```

One another useful function of dendrogram is cut the columns/rows into groups.
You can specify `k` or `h`, which works the same with `cutree`.

```{r}
ggheat(small_mat) + htanno_dendro(position = "top", k = 3)
```

`htanno_dendro` will create a default `node` and `edge` data for the ggplot. See
`?dendrogram_data` for details. `node` data have been set as the default data of
the ggplot object, and `edge` data has been added into the `geom_segment` layer
which drawing the dendrogram tree, so you cannot control the dendrogram by geom
function, but you can control all of the behaviour in `htanno_dendro`. The
default mapping can be controled via `mapping` argument, which is the same with
`ggheat` and all other annotation function. One useful variable in the data is
`branch` column, which corresponding to the `cutree` result.

```{r}
ggheat(small_mat) + htanno_dendro(aes(color = branch), position = "top", k = 3)
```

All annotation functions include a `set_context` argument that determines
whether to set the active component when adding the annotation. This argument
requires a boolean value of length `2`. If a scalar value is provided, it will
be recycled.  The first value determines whether the active component is set to
the annotation position. The second value will determine whether the active
component of annotation list in this position should be set as current
annotation. In the code mentioned above, the active component of the heatmap is
set to the `top` and the active component of the top annotation list will be set
to the dendrogram plot since `htanno_dendro` has a default `set_context` value
of `TRUE`. So you can easily add more `geoms` with the default `node` data.

```{r}
ggheat(small_mat) +
  htanno_dendro(aes(color = branch), position = "top", k = 3) +
  geom_point(aes(y = y))
```

By default, all annotation functions automatically create a default mapping for
the axis parallel to the heatmap. Therefore, when utilizing these functions, you
only need to provide the mapping for the other axis. 

As mentioned above, `facet_*` function won't work, since the internal will
always use the row/column groups controlled by heatmap annotation, but you can
also provide `facet_grid` or `facet_null` (if no panels) to control the
non-layout arguments, like `labeller`.

```{r}
ggheat(small_mat) +
  htanno_dendro(aes(color = branch), position = "top", k = 3) +
  geom_point(aes(y = y)) +
  facet_grid(labeller = labeller(.panel = function(x) letters[as.integer(x)]))
```

Each annotation function includes a `data` argument to specify the annotation
data. This argument can accept a matrix, a data frame, or even a simple vector
that will be converted into a one-column matrix. If the `data` argument is set to
`NULL`, the function will use the heatmap matrix, as shown in the previous
examples. Additionally, the data argument can also accept a function(purrr-like
lambda is also okay), to manipulate the heatmap matrix. It is important to note
that all annotations consider the rows as the observations. It means the `NROW`
function must return the same number as the heatmap parallel axis. So for column
annotation, the heatmap will be transposed before using (If `data` is a
function, it will be applied with the transposed matrix).

Since `htanno_dendro` has set the active context for the `top` position, we can
add more annotations without specifying the position arguments.

```{r}
ggheat(small_mat) +
  htanno_dendro(aes(color = branch), position = "top", k = 3) +
  # since the heatmap matrix will be transposed before using, we should to
  # rowSums to calculate the column sums of the original matrix
  gganno(data = rowSums) +
  # again: the default mapping will be made for the parallel axis of heatmap
  geom_bar(aes(y = value, fill = .panel), stat = "identity") +
  scale_fill_brewer(palette = "Dark2")
```

## `gganno`

`gganno` is similar to `ggheat` and `ggplot` in that it initializes a `ggplot`
data and `mapping`. Before drawing, specific operations are performed to prepare
for the visualization. The matrix (including a simple vector) is converted into
a long-format data frame, similar to the process utilized in ggheat. However,
the data frame does not contain .row_panel or .column_panel columns, as
annotations can only have one facet axis. In the case where the input data is
already a data frame, three additional columns-`.row_names`, `.row_index`, and
`.panel`—are added to the data frame. 

Let's add more annotations to different position. Since we want to add
annotation into another position, we must specify the `position` argument or we
can add an `active` object (we also provide `activate` and `deactivate`
for function call foramt).

```{r}
ggheat(small_mat) +
  scale_fill_viridis_c(guide = "none") +
  htanno_dendro(aes(color = branch), position = "top", k = 3) +
  gganno(data = rowSums) +
  geom_bar(aes(y = value, fill = .panel), stat = "identity") +
  scale_fill_brewer(palette = "Dark2") +
  active("left") +
  gganno(aes(x = value), data = rowSums) +
  geom_bar(
    aes(y = .y, fill = factor(.y)),
    stat = "identity",
    orientation = "y"
  ) +
  scale_fill_brewer(palette = "Set1", guide = "none") +
  scale_x_reverse()
```

For convenient, we also provide `gganno_*` function for 4 positions.
```{r}
ggheat(small_mat) +
  gganno_top(aes(y = value), data = rowSums) +
  geom_bar(stat = "identity", aes(fill = factor(.panel))) +
  scale_fill_brewer(name = NULL, palette = "Dark2", guide = "none") +
  gganno_left(aes(x = value), data = rowSums) +
  geom_bar(
    aes(y = .y, fill = factor(.y)),
    stat = "identity",
    orientation = "y"
  ) +
  scale_fill_brewer(palette = "Set1") +
  scale_x_reverse()
```

## `htanno_group`
Another `htanno_group` class provides convenient way to split heatmap
`rows/columns` into different panels with a group vairables. It won't draw
anything.

```{r}
ggheat(small_mat) +
  htanno_group(
    sample(letters[1:3], nrow(small_mat), replace = TRUE),
    position = "left"
  )
```

You can combine `htanno_dendro` and `htanno_group` in different axis.
```{r}
ggheat(small_mat) +
  htanno_dendro(aes(color = branch), position = "top", k = 3) +
  htanno_group(
    sample(letters[1:3], nrow(small_mat), replace = TRUE),
    position = "left"
  ) +
  gganno_left(aes(x = value), data = rowSums) +
  geom_bar(
    aes(y = .y, fill = factor(.y)),
    stat = "identity",
    orientation = "y"
  ) +
  scale_fill_brewer(palette = "Set1", guide = "none") +
  scale_x_reverse()
```

But you cannot mix them in the same axis.
```{r error=TRUE}
ggheat(small_mat) +
  htanno_dendro(aes(color = branch), position = "top", k = 3) +
  htanno_group(sample(letters[1:3], ncol(small_mat), replace = TRUE))
```

I'll add support of sub-clustering for `htanno_dendro`.

## Control size
The different ggplot objects were combined using the `patchwork` package.
Internally, the `ggheat_build` function was used to construct and merge all the
`ggplot` objects into a single `patchwork` object. 

`ggheat` has `width` and `height` to control the relative (you can also provide
a unit object) width and height. All annotation function have a `size` argument
to control the relative width (left and right annotation) or height (top and
bottom annotation).

```{r}
pp <- ggheat(small_mat) +
  scale_fill_viridis_c() +
  gganno_top(data = rowSums, size = unit(10, "mm")) +
  geom_bar(aes(y = value, fill = .x), stat = "identity") +
  gganno_left(aes(x = value), data = rowSums, size = 0.5) +
  geom_bar(
    aes(y = .y, fill = factor(.y)),
    stat = "identity",
    orientation = "y"
  ) +
  scale_fill_brewer(palette = "Set1", guide = "none") +
  scale_x_reverse()
pp
```

Internally, the `ggheat_build` function was used to construct and merge all the
`ggplot` objects into a single `patchwork` object. But usually there is no need
to manually perform this combination since much of the arguments can be
controlled within the `ggheat` function and the annotation functions.
```{r}
class(ggheat_build(pp))
```

## Session information
```{r}
sessionInfo()
```
