---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

## Installation

You can install the development version of `ggalign` from [GitHub](https://github.com/) with:

```{r, eval=FALSE}
# install.packages("remotes")
remotes::install_github("Yunuuuu/ggalign")
```

This package now depends on some un-merged features of patchwork. Please see
<https://github.com/thomasp85/patchwork/pull/373>. If you want to try it, you
should install patchwork with
`pak::pkg_install(Yunuuuu/patchwork@align_axis_title)`.

The documentation for `ggalign` is currently in progress.

# ggalign
This package extends ggplot2 and provides numerous benefits for organizing and
arranging plots. It is specifically designed to align a specific axis of
multiple ggplot objects in a consistent order. This functionality is
particularly useful for plots that require manipulation of data order. A common
plot combination that can be effectively organized using this package includes a
dendrogram and a heatmap.

## Features
`ggalign` provides two layout to arrange ggplot objects:

  - Heatmap layout: Arrange ggplot into a Heatmap. See
  `vignette("layout-heatmap")` for details.

  - Stack layout: Arrange ggplot vertically or horizontally. See
  `vignette("layout-stack")` for details.

`ggalign` provides `Align` objects to reordering or set panels for layout, in
addition, `Align` objects can also add plot into the layout:

  - `align_group()`: Group layout axis into panel
  - `align_kmeans()`: Group layout observations by kmeans
  - `align_reorder()`: Reorder layout observations
  - `align_dendro()`: Reorder or Group layout based on Hierarchical Clustering
  - `align_gg()`/`ggalign`: Create ggplot object in the layout
  - `align_panel()`/`ggalign`: Create ggplot object based on the layout panel data.

See `vignette("align")` for details.

## Examples


## Compare with ComplexHeamtap
