---
title: "align"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{align}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ggalign)
```

# `align_*`
The `ggalign` package provides a range of `align_*` functions that enable
control over the layout. Specifically, these functions allow for the
manipulation of axis order in the layout, as well as partitioning along a single
axis into multiple panel. Additionally, some of these functions have the
capability to add plots. 

Currently, there are four primary `align_*` functions: `align_group`,
`align_reorder`, `align_kmeans` and `align_dendro`.

```{r setup_data}
set.seed(123)
small_mat <- matrix(rnorm(81), nrow = 9)
rownames(small_mat) <- paste0("row", seq_len(nrow(small_mat)))
colnames(small_mat) <- paste0("column", seq_len(ncol(small_mat)))
```

## `align_group`

`align_group` just group along a layout axis into different panel. It won't add
any plot area.

```{r align_group_top}
ggheatmap(small_mat) +
  hmanno("t") +
  align_group(sample(letters[1:4], ncol(small_mat), replace = TRUE))
```

By default, the facet strip text will be removed, you can override the behaviour
with `theme(strip.text = element_text())`. Because `align_group` has no plot
area, you can only add the panel title in the heatmap plot.
```{r align_group_left}
ggheatmap(small_mat) +
  theme(strip.text = element_text()) +
  hmanno("l") +
  align_group(sample(letters[1:4], nrow(small_mat), replace = TRUE))
```

## `align_reorder`
`align_reorder` reorder the layout rows/columns based on a summary function. It
won't add any plot area. 
```{r}
ggheatmap(small_mat) +
  hmanno("l") +
  align_reorder(rowMeans)
```

Some `align_*` functions accept a `data` argument. This argument can be a
matrix, a data frame, or even a simple vector that will be converted into a
one-column matrix. If the `data` argument is set to `NULL`, the function will
use the layout data, as shown in the previous example. Additionally, the `data`
argument can also accept a function (purrr-like lambda is also okay), which will
be applied with the layout data. 

> It is important to note that all `align_*` function consider the rows as the
observations. It means the `NROW` function must return the same number as the
layout parallel axis. So for heatmap column annotation, the heatmap matrix will
be transposed before using (If `data` is a function, it will be applied with the
transposed matrix).

So even for top and bottom annotation, we can also use the `rowMeans` to
calculate the mean value across all columns.

```{r}
ggheatmap(small_mat) +
  hmanno("t") +
  align_reorder(rowMeans)
```

## `align_kmeans`
`align_kmeans` group heatmap rows/columns by kmeans. It won't add any plot area. 
```{r}
ggheatmap(small_mat) +
  hmanno("t") +
  align_kmeans(3L)
```

It is important to note that both `align_group` and `align_kmeans` cannot do
sub-groups. It means you cannot use them when there has been any groups exist.

```{r error=TRUE}
ggheatmap(small_mat) +
  hmanno("t") +
  align_group(sample(letters[1:4], ncol(small_mat), replace = TRUE)) +
  align_kmeans(3L)
```

```{r error=TRUE}
ggheatmap(small_mat) +
  hmanno("t") +
  align_kmeans(3L) +
  align_group(sample(letters[1:4], ncol(small_mat), replace = TRUE))
```

### align_dendro
`align_dendro` class the addition of a dendrogram near the heatmap and and it
can also reorder the heatmap. This is primarily useful when working with
heatmap plots.

```{r}
ggheatmap(small_mat) +
  hmanno("t") +
  align_dendro()
```

You can use the `distance` and `method` argument to control the dendrogram
builind proce.

```{r}
ggheatmap(small_mat) +
  hmanno("t") +
  align_dendro(method = "ward.D2")
```

One useful function of dendrogram is to cut the columns/rows into groups. You
can specify `k` or `h`, which works the same with `cutree`.

```{r}
ggheatmap(small_mat) +
  hmanno("t") +
  align_dendro(k = 3L)
```

In contrast to `align_group`, `align_kmeans`, and `align_reorder`,
`align_dendro` is capable of drawing plot components. Therefore, it has a
default `set_context` value of `TRUE`, which indicates whether to set the active
context to current `Align` object. Consequently, any ggplot elements added
thereafter will be included in this plot area.

```{r}
ggheatmap(small_mat) +
  hmanno("t") +
  align_dendro() +
  geom_point(aes(y = y))
```

`align_dendro` will create a default `node` data for the ggplot. See
`?dendrogram_data` for details. In addition, `edge` data has been added into the
`geom_segment` layer directly which was used to draw the dendrogram tree. One
useful variable in both `node` and edge `data` is `branch` column, which
corresponding to the `cutree` result.

```{r}
ggheatmap(small_mat) +
  hmanno("t") +
  align_dendro(aes(color = branch), k = 3) +
  geom_point(aes(color = branch, y = y))
```

`align_dendro` is also capable of performing clustering between groups. This
means that you can use it even if there are already existing groups present in
the layout.

```{r}
column_groups <- sample(letters[1:3], ncol(small_mat), replace = TRUE)
ggheatmap(small_mat) +
  hmanno("t") +
  align_group(column_groups) +
  align_dendro(aes(color = branch))
```

You can reorder the groups by setting `reorder_group = TRUE`.
```{r}
ggheatmap(small_mat) +
  hmanno("t") +
  align_group(column_groups) +
  align_dendro(aes(color = branch), reorder_group = TRUE)
```

You can see the difference.
```{r}
ggheatmap(small_mat) +
  hmanno("t") +
  align_group(column_groups) +
  align_dendro(aes(color = branch), reorder_group = TRUE) +
  hmanno("b") +
  align_dendro(aes(color = branch), reorder_group = FALSE)
```

It is important to understand that when `reorder_group = FALSE` is used, the
reordering of the heatmap occurs within each group. As long as the ordering
within each group remains the same, these two dendrograms can be placed on the
same axis of the heatmap. This is why adding a dendrogram with `reorder_group =
FALSE` behind a dendrogram with `reorder_group = TRUE` is acceptable, as the
second `align_dendro` will follow the group ordering established by the first
one. However, it is not possible to add a dendrogram with `reorder_group = TRUE`
behind a dendrogram with `reorder_group = FALSE` because the second dendrogram
would not be able to conform to the grouping order followed by the first
dendrogram.

```{r error=TRUE}
ggheatmap(small_mat) +
  hmanno("t") +
  align_group(column_groups) +
  align_dendro(aes(color = branch), reorder_group = FALSE) +
  hmanno("b") +
  align_dendro(aes(color = branch), reorder_group = TRUE)
```

> We always prevent users from reordering the layout axis twice.

## `align_gg`
`align_gg` is similar to `ggplot` in that it initializes a `ggplot` data and
`mapping`. The data input can be a matrix, a data frame, or a simple vector that
will be converted into a one-column matrix, and can inherit from the layout
data.

The internal will always use a default mapping for the parallel axis with layout
(`aes(y = .data$.y)` or `aes(x = .data$.x)`).

For ggplot usage, matrix (including a simple vector) data is converted into a
long-format data frame, similar to the process utilized in `ggheatmap`. But
note that the long-format data frame does not contain `.row_panel` or
`.column_panel` column, as `align_gg` can only have one facet axis. In the
case where the input data is already a data frame, three additional
columns-(`.row_names`, `.row_index`, and `.panel`)—are added to the data
frame.

The data in the underlying `ggplot` object contains following columns:

  - `.panel`: the panel for current annotation

  - `.row_names` and `.row_index`: the row names and row index of the original
    matrix or data frame.

  - `.column_names` and `.column_index`: the row and column index of the
    original matrix (only applicable if `data` is a `matrix`).

  - `.x` or `.y`: the `x` or `y` coordinates

  - `value`: the actual matrix value (only applicable if `data` is a
    `matrix`).

You can also use `ggalign` function which is an alias of `align_gg`.

```{r}
ggheatmap(small_mat) +
  scale_fill_viridis_c(guide = "none") +
  hmanno("t") +
  align_dendro(aes(color = branch), k = 3) +
  ggalign(data = rowSums) +
  geom_bar(aes(y = value, fill = .panel), stat = "identity") +
  scale_fill_brewer(palette = "Dark2") +
  hmanno("l") +
  ggalign(aes(x = value), data = rowSums) +
  geom_bar(
    aes(y = .y, fill = factor(.y)),
    stat = "identity",
    orientation = "y"
  ) +
  scale_fill_brewer(palette = "Set1") +
  scale_x_reverse()
```

## Session information
```{r}
sessionInfo()
```
