---
title: "Heatmap Layout"
output: 
  bookdown::html_document2:
    base_format: rmarkdown::html_vignette
    number_sections: no
    fig_caption: yes
    fig_width: 7
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Heatmap Layout}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
pkgdown:
  as_is: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The `heatmap_layout()`/`ggheatmap()` function provides a powerful way to create
customizable heatmaps in R using ggplot2. This vignette will guide you through
its usage.

## Input Data
The data input can be a numeric or character vector, a data frame, and any other
data which can be converted into a matrix.

```{r setup_data}
set.seed(123)
small_mat <- matrix(rnorm(81), nrow = 9)
rownames(small_mat) <- paste0("row", seq_len(nrow(small_mat)))
colnames(small_mat) <- paste0("column", seq_len(ncol(small_mat)))
```

```{r matrix}
library(ggalign)
ggheatmap(small_mat)
```

## heatmap body
For `ggplot2` usage, the matrix input will be converted into a long formated
data frame when drawing. The default mapping will use `aes(.data$.x, .data$.y)`,
but can be controlled using `mapping` argument. The data in the underlying
ggplot object contains following columns:

  - `.xpanel` and `.ypanel`: the column and row panel

  - `.x` and `.y`: the `x` and `y` coordinates

  - `.row_names` and `.column_names`: A factor of the row and column names of
    the original matrix (only applicable when names exist).

  - `.row_index` and `.column_index`: the row and column index of the original
    matrix.

  - `value`: the actual matrix value.

You can treat the `heatmap_layout()` object as a normal ggplot2 object with a
default `mapping` and `data`. You can add `ggplot2` elements as usual.
```{r}
ggheatmap(small_mat) + geom_point() + scale_fill_viridis_c()
```

By default, a heatmap layer is added. If you set filling = FALSE, a blank
heatmap will be drawn, allowing for customized filling geoms. For larger
matrices, you may want to use a raster layer for faster rendering. In such
cases, packages like [ggrastr](https://github.com/VPetukhov/ggrastr) or
[ggfx](https://ggfx.data-imaginist.com/reference/with_raster.html) are helpful.
Additionally, [ggfx](https://ggfx.data-imaginist.com/reference/with_raster.html)
offers many image filters that can be applied to ggplot2 layers. 
```{r eval=rlang::is_installed("ragg")}
ggheatmap(small_mat, filling = FALSE) +
  ggrastr::rasterise(geom_tile(aes(fill = value)), dev = "ragg")
```

Indeed, we have add `ggrastr::rasterise()` method for both `ggheatmap()` and
`ggstack()` objects, you can directly call `ggrastr::rasterise()` to the
`ggheatmap()` which will rasterise all plots in the layout.
```{r eval=rlang::is_installed("ragg")}
ggrastr::rasterise(ggheatmap(small_mat), dev = "ragg")
```

## heatmap annotations
A heatmap typically has two observation axes, you can reorder the observations,
or provide additional information for them. An annotation is a `stack_layout()`
object internally, capable of holding multiple plots, and can be placed at the
`top`, `left`, `bottom`, or `right`.

Note that `ggalign` use the concept of `"number of observations"` in the
[vctrs](https://vctrs.r-lib.org/reference/vec_size.html) package or `NROW()`
function. When aligning the observations, you must ensure the number of
observations is equal. For column annotations, the layout data will be
transposed before use. This is necessary because column annotations use heatmap
columns as observations, but we need rows. 

By default, `heatmap_layout()`/`ggheatmap()` does not initialize an active
context, so all additions are placed within the heatmap body. we can use
`hmanno()` to set the active context, directing all subsequent additions to this
position. The active context allows for custom layout adjustments and the
addition of various plot types. 

In the following example, `align_kmeans()` is used to group the columns into three
panels. It doesn't matter if this is added to the top or bottom since it won't
add a plot area: 

```{r}
ggheatmap(small_mat) +
  hmanno("t") +
  align_kmeans(3L)
```

We can add any `align_*()` function to the annotation. For more details on
`align_*()` functions, refer to `vignette("layout-customize")` and
`vignette("layout-plot")`.

```{r}
ggheatmap(small_mat) +
  theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
  hmanno("r") +
  align_dendro(k = 3L) +
  geom_point(aes(color = factor(branch)))
```

In this example:

 - `hmanno("r")` change the active context to the right of the heatmap.
 - `align_dendro(k = 3L)` adds a dendrogram to this right-side annotation
 context and sets itself as the active plot in the annotation stack. 
 - `geom_point(aes(color = factor(branch)))` is then added to this active plot
 within the annotation stack, here, it means the `align_dendro()` plot.

## plot size
### heatmap body size
You can specify the heatmap body relative sizes (`width`/`height`) in the
`ggheatmap()` function.
```{r}
ggheatmap(small_mat, height = 2) +
  scale_fill_viridis_c() +
  hmanno("t") +
  align_dendro()
```

Alternatively, the `hmanno()` function allows you to control the heatmap body
sizes when `position` is `NULL`. 
```{r}
ggheatmap(small_mat) +
  hmanno(position = NULL, height = 2) +
  scale_fill_viridis_c() +
  hmanno("t") +
  align_dendro()
```

### heatmap annotation total size
The `hmanno()` function allows you to control the total annotation stack size
when `position` is a `string`. The `size` argument controls the relative width
(for left and right annotations) or height (for top and bottom annotations) of
the whole annotation stack. 

```{r}
ggheatmap(small_mat) +
  scale_fill_viridis_c() +
  hmanno("t", size = 1) +
  align_dendro()
```

We can use `unit()` to define the size.
```{r}
ggheatmap(small_mat) +
  scale_fill_viridis_c() +
  hmanno("t", size = unit(30, "mm")) +
  align_dendro()
```

### single annotation plot size
All `align_*()` functions that add plots in the annotation stack have a `size`
argument to control the relative width (for left and right annotations) or
height (for top and bottom annotations) of the single plot in the annotation
stack. 
```{r}
ggheatmap(small_mat) +
  scale_fill_viridis_c() +
  hmanno("l", size = 0.2) +
  ggalign(data = rowSums, aes(x = value), size = unit(10, "mm")) +
  geom_bar(
    aes(y = .y, fill = factor(.y)),
    stat = "identity",
    orientation = "y"
  ) +
  scale_fill_brewer(palette = "Set1", guide = "none")
```

## guide legends
By default, `ggheatmap()` will collect all guide legends on the side from which
they originate. 

```{r fig.dim = c(12, 12)}
heatmap_with_four_side_guides <- ggheatmap(small_mat) +
  scale_fill_viridis_c(name = "I'm from heatmap body") +
  theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
  hmanno("t") +
  align_dendro(aes(color = branch), k = 3L) +
  scale_color_brewer(
    name = "I'm from top annotation", palette = "Dark2",
    guide = guide_legend(position = "right")
  ) +
  hmanno("l") +
  align_dendro(aes(color = branch), k = 3L) +
  scale_color_brewer(
    name = "I'm from left annotation", palette = "Dark2",
    guide = guide_legend(position = "top")
  ) +
  hmanno("b") +
  align_dendro(aes(color = branch), k = 3L) +
  scale_color_brewer(
    name = "I'm from bottom annotation", palette = "Dark2",
    guide = guide_legend(position = "left")
  ) +
  hmanno("r") +
  align_dendro(aes(color = branch), k = 3L) +
  scale_color_brewer(
    name = "I'm from right annotation", palette = "Dark2",
    guide = guide_legend(position = "bottom")
  ) &
  theme(plot.margin = margin())
heatmap_with_four_side_guides
```

This package provides the `guides` and `free_guides` arguments for precise control
over guide legends.

### which position of guide legends should be collected in the layout?
The `guides` argument controls which positions of guide legends should be
collected in the layout. They will be collected to their original side. 

#### heatmap layout
For a heatmap layout, it has five plots (heatmap body, four sides of the nested
heatmap annotation stack). When position is `NULL`, `hmanno()` set the `guides`
for the heatmap layout.
```{r heatmap-guides-tb, fig.dim = c(12, 12), fig.cap="Heatmap collect top and bottom guides only"}
heatmap_with_four_side_guides +
  # the heatmap layout only collect guide legends in the top and bottom
  hmanno(guides = "tb")
```

In this way, the guide legends in the left and right position won't be
collected.

#### heatmap annotation stack layout
In the nested layouts, the top-level layout won't collect guide legends from
plots within the nested layout unless the nested layout collects them first.
Guide legends in the heatmap layout will only be collected from the annotation
stack if the stack layout collects its own guide legends first.

When `position` is a string, `hmanno()` set the `guides` for the heatmap
annotation stack. By default, it'll inherit from the parent layout (heatmap
layout)'s `guides` argument.
```{r fig.dim = c(12, 12)}
heatmap_with_four_side_guides +
  # the left annotation won't collecte the guide legends
  hmanno("l", guides = NULL)
```

>Note: Guide legends in the heatmap layout will only be collected from the
annotation stack if the stack layout collects its own guide legends first.

In this example, the guide legends from the left annotation won't be collected
by the heatmap layout since the left annotation doesn't collect the guide
legends. 

### which position of guide legends should be collected for a single plot?
The `free_guides` argument allows you to override the `guides` argument for a
single plot.

#### heatmap body
When `position` is `NULL`, `hmanno()` set the `free_guides` for the heatmap
body.
```{r fig.dim = c(12, 12)}
heatmap_with_four_side_guides +
  # the heatmap layout only collect guide legends in the top and bottom
  hmanno(guides = "tb") +
  # we override the layout `guides` argument and collect the right guide
  # legend for the heatmap body
  hmanno(free_guides = "r")
```

In this example, the heatmap body right side legend won't be collected by
default in the layout. We use `free_guides` to override the `guides` argument
for the heatmap body plot.

#### heatmap annotation stack
In the heamtap layout, each annotation stack is a nested stack plot, we can also
use `free_guides` to override the `guides` argument in the heatmap layout for
the annotation stack in `hmanno()`.

```{r fig.dim = c(12, 12)}
heatmap_with_four_side_guides +
  # the heatmap layout only collect guide legends in the top and bottom
  hmanno(guides = "tb") +
  # we collect the right guide legend for the right annotation in the heatmap layout
  hmanno("t", free_guides = "r") +
  # we collect the left guide legend for the bottom annotation in the heatmap layout
  hmanno("b", free_guides = "l")
```

#### single plot in the annotation stack
All `align_*()` functions which add plot include a `free_guides` argument to
override the `guides` setting for the plot in the annotation stack layout.
```{r}
ggheatmap(small_mat) +
  scale_fill_viridis_c() +
  theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
  hmanno("t") +
  # we won't collect any guide legends from the dendrogram plot
  align_dendro(aes(color = branch), k = 3L, free_guides = NULL) +
  align_dendro(aes(color = branch), k = 3L, free_guides = NULL) &
  scale_color_brewer(
    palette = "Dark2",
    guide = guide_legend(position = "bottom")
  )
```

## free_spaces
By default, `ggheatmap()` will align all elements of the plot, which can
sometimes lead to unwanted spacing. Consider the following example:

```{r}
ggheatmap(small_mat) +
  scale_fill_viridis_c() +
  hmanno("t", size = unit(30, "mm")) +
  align_dendro() +
  scale_y_continuous(
    expand = expansion(),
    labels = ~ paste("very very long labels", .x)
  ) +
  hmanno("l", unit(20, "mm")) +
  align_dendro()
```

Both `free_spaces` and `free_labs` operate on individual plots, while layout
values set global parameters for the entire layout. This means that `align_*`
functions will inherit these values, but you can override them within the
`align_*` functions.

- **If `position` is `NULL`**: These arguments control the behavior of the heatmap 
  body and set the global values for the heatmap layout. Setting to `waiver()` 
  inherits from the parent layout.

- **If `position` is a string**: These arguments set the global values for the 
  annotation stack layout. If set to `waiver()`, it inherits from specific
  heatmap layout axes:
  - For top and bottom annotations, it inherits from the left ("l") and right ("r") 
    axes.
  - For left and right annotations, it inherits from the top ("t") and bottom ("b") 
    axes.

In this case, the left annotation stack is positioned far from the heatmap body
due to the wide axis labels in the top annotation stack. This occurs because the
top annotation stack is aligned with the heatmap body. To fix this, you can
remove the left borders around the panel of the top annotation stack by setting
`free_spaces = "l"`.

```{r}
ggheatmap(small_mat) +
  scale_fill_viridis_c() +
  hmanno("t", size = unit(30, "mm"), free_spaces = "l") +
  align_dendro() +
  scale_y_continuous(
    expand = expansion(),
    labels = ~ paste("very very long labels", .x)
  ) +
  hmanno("l", unit(20, "mm")) +
  align_dendro()
```

In the Figure \@ref(fig:heatmap-guides-tb) we can also use the `free_spaces` to
make a compact heatmap. Here, we remove the right border spaces of the top
annotation stack, and the left border spaces of the bottom annotation stack.
```{r fig.dim = c(12, 12)}
heatmap_with_four_side_guides +
  hmanno(guides = "tb") + # we set the `guides` argument here
  hmanno("t", free_spaces = "r") + # we remove the right border spaces
  hmanno("b", free_spaces = "l") # we remove the left border spaces
```

`align_*()` functions include the `free_spaces` argument, which, by default,
inherits from the annotation stack. This allows you to control the alignment
for each plot individually.

```{r}
ggheatmap(small_mat) +
  scale_fill_viridis_c() +
  hmanno("t", size = unit(30, "mm")) +
  align_dendro(free_labs = "l") +
  scale_y_continuous(
    expand = expansion(),
    labels = ~ paste("very very long labels", .x)
  ) +
  hmanno("l", unit(20, "mm")) +
  align_dendro()
```

## free_labs
By default, `ggheatmap()` won't align the axis titles.
```{r}
ggheatmap(small_mat) +
  scale_fill_viridis_c() +
  ylab("Heatmap title") +
  hmanno("t", size = unit(30, "mm")) +
  align_dendro() +
  ylab("Annotation title")
```

To align all axis titles, you can set `free_labs = NULL` in the `hmanno()`
function. Alternatively, A single string containing one or more of axis
positions ("t", "l", "b", "r") to indicate which axis titles should be free from
alignment.

```{r}
ggheatmap(small_mat) +
  hmanno(free_labs = NULL) +
  scale_fill_viridis_c() +
  ylab("Heatmap title") +
  hmanno("t", size = unit(30, "mm")) +
  align_dendro() +
  ylab("Annotation title")
```

The same with `free_spaces` argument, you can control the `free_labs` argument
for the individual plot in `align_*()` function.

## Session information
```{r}
sessionInfo()
```
