---
title: "Heatmap Layout"
output: 
  bookdown::html_document2:
    base_format: rmarkdown::html_vignette
    number_sections: no
    fig_caption: yes
    fig_width: 7
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Heatmap Layout}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
pkgdown:
  as_is: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The `heatmap_layout()`/`ggheatmap()` function provides a powerful way to create
customizable heatmaps in R using ggplot2. This vignette will guide you through
its usage.

# input data
The data input can be a numeric or character vector, a data frame, and any other
data which can be converted into a matrix.

```{r setup_data}
set.seed(123)
small_mat <- matrix(rnorm(81), nrow = 9)
rownames(small_mat) <- paste0("row", seq_len(nrow(small_mat)))
colnames(small_mat) <- paste0("column", seq_len(ncol(small_mat)))
```

```{r matrix}
library(ggalign)
ggheatmap(small_mat)
```

# heatmap body
For `ggplot2` usage, the matrix input will be converted into a long formated
data frame when drawing. The default mapping will use `aes(.data$.x, .data$.y)`,
but can be controlled using `mapping` argument. The data in the underlying
ggplot object contains following columns:

  - `.xpanel` and `.ypanel`: the column and row panel

  - `.x` and `.y`: the `x` and `y` coordinates

  - `.row_names` and `.column_names`: A factor of the row and column names of
    the original matrix (only applicable when names exist).

  - `.row_index` and `.column_index`: the row and column index of the original
    matrix.

  - `value`: the actual matrix value.

You can treat the `heatmap_layout()` object as a standard ggplot2 object with
default mapping and data. This means you can add ggplot2 layers or elements just
like in any ggplot object.
```{r}
ggheatmap(small_mat) + geom_point() + scale_fill_viridis_c()
```

By default, `ggheatmap()` adds a heatmap layer. If the matrix has more than
20,000 cells (`nrow * ncol > 20000`), `geom_raster()` will be used to render the
heatmap body for performance efficiency. For smaller matrices, `geom_tile()` is
used. 

If you set `filling = FALSE`, a blank heatmap will be drawn, allowing for
customized filling geoms. You can use external packages like
[ggrastr](https://github.com/VPetukhov/ggrastr) or
[ggfx](https://ggfx.data-imaginist.com/reference/with_raster.html) to rasterize
the heatmap body. Additionally,
[ggfx](https://ggfx.data-imaginist.com/reference/with_raster.html) offers many
image filters that can be applied to ggplot2 layers.
```{r eval=rlang::is_installed("ragg")}
ggheatmap(small_mat, filling = FALSE) +
  ggrastr::rasterise(geom_tile(aes(fill = value)), dev = "ragg")
```

You can rasterize all plots in the layout directly with `ggrastr::rasterise()`.
This method works for both `ggheatmap()` and `ggstack()` objects.
```{r eval=rlang::is_installed("ragg")}
ggrastr::rasterise(ggheatmap(small_mat), dev = "ragg")
```

# heatmap annotations
A heatmap typically has two observation axes, you can reorder the observations,
or provide additional information for them. An annotation is a `stack_layout()`
object internally, capable of holding multiple plots, and can be placed at the
`top`, `left`, `bottom`, or `right`.

Note that `ggalign` use the concept of `"number of observations"` in the
[vctrs](https://vctrs.r-lib.org/reference/vec_size.html) package or `NROW()`
function. When aligning the observations, you must ensure the number of
observations is equal. For column annotations, the layout data will be
transposed before use. This is necessary because column annotations use heatmap
columns as observations, but we need rows. 

By default, `heatmap_layout()`/`ggheatmap()` does not initialize an active
context, so all additions are placed within the heatmap body. we can use
`hmanno()` to set the active context, directing all subsequent additions to this
position. The active context allows for custom layout adjustments and the
addition of various plot types. 

In the following example, `align_kmeans()` is used to group the columns into three
panels. It doesn't matter if this is added to the top or bottom since it won't
add a plot area: 

```{r}
ggheatmap(small_mat) +
  hmanno("t") +
  align_kmeans(3L)
```

We can add any `align_*()` function to the annotation. For more details on
`align_*()` functions, refer to `vignette("layout-customize")` and
`vignette("layout-plot")`.

```{r}
ggheatmap(small_mat) +
  theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
  hmanno("r") +
  align_dendro(k = 3L) +
  geom_point(aes(color = factor(branch)))
```

In this example:

 - `hmanno("r")` change the active context to the right of the heatmap.
 - `align_dendro(k = 3L)` adds a dendrogram to this right-side annotation
 context and sets itself as the active plot in the annotation stack. 
 - `geom_point(aes(color = factor(branch)))` is then added to this active plot
 within the annotation stack, here, it means the `align_dendro()` plot.

# Plot Size
## Heatmap Body Size
You can specify the relative sizes of the heatmap body using the `.width` and
`.height` arguments in the `ggheatmap()` function. The dot prefix helps avoid
conflicts with arguments from `geom_tile()`/`geom_raster()`. 
```{r}
ggheatmap(small_mat, .height = 2) +
  scale_fill_viridis_c() +
  hmanno("t") +
  align_dendro()
```

Alternatively, the `hmanno()` function allows you to control the heatmap body
sizes when `position` is `NULL`. 
```{r}
ggheatmap(small_mat) +
  hmanno(position = NULL, height = 2) +
  scale_fill_viridis_c() +
  hmanno("t") +
  align_dendro()
```

## Annotation Stack Size
The `hmanno()` function allows you to control the total annotation stack size
when `position` is a `string`. The `size` argument controls the relative width
(for left and right annotations) or height (for top and bottom annotations) of
the whole annotation stack. 

```{r}
ggheatmap(small_mat) +
  scale_fill_viridis_c() +
  hmanno("t", size = 1) +
  align_dendro()
```

You can also specify an exact size using `unit()`:
```{r}
ggheatmap(small_mat) +
  scale_fill_viridis_c() +
  hmanno("t", size = unit(30, "mm")) +
  align_dendro()
```

## Single Plot Size
All `align_*()` functions that add plots in the annotation stack have a `size`
argument to control the relative width (for left and right annotations) or
height (for top and bottom annotations) of the single plot in the annotation
stack. 
```{r}
ggheatmap(small_mat) +
  scale_fill_viridis_c() +
  hmanno("l", size = 0.2) +
  ggalign(data = rowSums, aes(x = value), size = unit(10, "mm")) +
  geom_bar(
    aes(y = .y, fill = factor(.y)),
    stat = "identity",
    orientation = "y"
  ) +
  scale_fill_brewer(palette = "Set1", guide = "none")
```

# Plot actions
The `plot_action()` function defines the behavior of plots within a layout. It
can be used in the `action` argument of layout functions like `hmanno()` or
`stack_active()` to set global actions for all plots in the layout.
Additionally, `plot_action()` can be applied directly to specific plots through
the `action` argument in the `align_*()` functions, or it can be added directly
to a plot.

## guides
By default, `ggheatmap()` will collect all guide legends on the side from which
they originate. 

```{r fig.dim = c(12, 12)}
heatmap_collect_all_guides <- ggheatmap(small_mat, .width = 2, .height = 2L) +
  scale_fill_viridis_c(name = "I'm from heatmap body") +
  theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
  hmanno("t") +
  align_dendro(aes(color = branch), k = 3L) +
  scale_color_brewer(
    name = "I'm from top annotation", palette = "Dark2",
    guide = guide_legend(position = "right")
  ) +
  hmanno("l") +
  align_dendro(aes(color = branch), k = 3L) +
  scale_color_brewer(
    name = "I'm from left annotation", palette = "Dark2",
    guide = guide_legend(position = "top", direction = "vertical")
  ) +
  hmanno("b") +
  align_dendro(aes(color = branch), k = 3L) +
  scale_color_brewer(
    name = "I'm from bottom annotation", palette = "Dark2",
    guide = guide_legend(position = "left")
  ) +
  hmanno("r") +
  align_dendro(aes(color = branch), k = 3L) +
  scale_color_brewer(
    name = "I'm from right annotation", palette = "Dark2",
    guide = guide_legend(position = "bottom", direction = "vertical")
  ) &
  theme(plot.margin = margin())
heatmap_collect_all_guides
```

Use the `guides` argument control which side of guide legends should be
gathered. In the following example, we'll collect the guide legends only on the
`top` and `bottom` sides: 
```{r fig.dim = c(12, 12)}
heatmap_collect_all_guides +
  # we only collect guides in the top and bottom side
  hmanno(action = plot_action(guides = "tb"))
```

You can also apply the `plot_action()` function directly to specific plots:
```{r fig.dim = c(12, 12)}
heatmap_collect_all_guides +
  # we only collect guides in the top and bottom side
  hmanno(action = plot_action(guides = "tb")) +
  # for the heatmap body, we also collect guide in the right side
  hmanno() +
  plot_action(guides = "r")
```

As previously mentioned, the annotation is represented as a `stack_layout()`
object. Guide legends within the annotation stack are first collected by the
`stack_layout()` itself and are then passed to the overall heatmap layout for
further collection.  See following example:
```{r fig.dim = c(12, 12)}
heatmap_collect_all_guides +
  # we only collect guides in the top and bottom side
  hmanno(action = plot_action(guides = "tb")) +
  # for the bottom annotation, we also collect guide in the left side
  hmanno("b") +
  plot_action(guides = "l")
```

Here, the guide legend is collected by the bottom annotation but will not be
collected by the heatmap layout since the heatmap layout only gathers guides
from the `top` and `bottom`.

To override the guide collection behavior of the heatmap annotation, you can use
the `free_guides` argument of the `hmanno()` function.
```{r fig.dim = c(12, 12)}
heatmap_collect_all_guides +
  # we only collect guides in the top and bottom side
  hmanno(action = plot_action(guides = "tb")) +
  # for the bottom annotation, we also collect guide in the left side
  hmanno("b", free_guides = "l") +
  plot_action(guides = "l")
```

>Note: The heatmap layout will only collect guide legends from the annotation
stack if the stack layout collects its own guides first. 

## free_spaces
By default, `ggheatmap()` will align all elements of the plot, which can
sometimes lead to unwanted spacing. Consider the following example:

```{r}
ggheatmap(small_mat) +
  scale_fill_viridis_c() +
  hmanno("t", size = unit(30, "mm")) +
  align_dendro() +
  scale_y_continuous(
    expand = expansion(),
    labels = ~ paste("very very long labels", .x)
  ) +
  hmanno("l", unit(20, "mm")) +
  align_dendro()
```

In this case, the left annotation stack is positioned far from the heatmap body
due to the wide axis labels in the top annotation stack. This occurs because the
top annotation stack is aligned with the heatmap body. To fix this, you can
remove the left borders around the panel of the top annotation stack by setting
`free_spaces = "l"`.

```{r}
ggheatmap(small_mat) +
  scale_fill_viridis_c() +
  hmanno("t", size = unit(30, "mm"), action = plot_action(free_spaces = "l")) +
  align_dendro() +
  scale_y_continuous(
    expand = expansion(),
    labels = ~ paste("very long labels", .x)
  ) +
  hmanno("l", unit(20, "mm")) +
  align_dendro()
```

In the `ggheatmap()` layout, the behavior of `free_spaces` and `free_labs`
arguments is distinct from other arguments in `plot_action()`:

- **If `position` is `NULL`**: These arguments control the behavior of the heatmap 
  body and set the global values for the heatmap layout. Setting to `waiver()` 
  inherits from the parent layout.

- **If `position` is a string**: These arguments set the global values for the 
  annotation stack layout. If set to `waiver()`, it inherits from specific
  heatmap layout axes:
  - For top and bottom annotations, it inherits from the left ("l") and right ("r") 
    axes.
  - For left and right annotations, it inherits from the top ("t") and bottom ("b") 
    axes.

## free_labs
By default, `ggheatmap()` won't align the axis titles.
```{r}
ggheatmap(small_mat) +
  scale_fill_viridis_c() +
  ylab("Heatmap title") +
  hmanno("t", size = unit(30, "mm")) +
  align_dendro() +
  ylab("Annotation title")
```

To align axis titles, you can set `free_labs = NULL`. Alternatively, A single
string containing one or more of axis positions ("t", "l", "b", "r") to indicate
which axis titles should be free from alignment.

```{r}
ggheatmap(small_mat, action = plot_action(free_labs = NULL)) +
  scale_fill_viridis_c() +
  ylab("Heatmap title") +
  hmanno("t", size = unit(30, "mm")) +
  align_dendro() +
  ylab("Annotation title")
```

# Session information
```{r}
sessionInfo()
```
