---
title: "Introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# ggalign
This package extends ggplot2 and provides numerous benefits for organizing and
arranging plots. It is specifically designed to align a specific axis of
multiple ggplot objects in a consistent order.  This functionality is
particularly useful for plots that require manipulation of data order. A common
plot combination that can be effectively organized using this package includes a
dendrogram and a heatmap.

```{r setup}
library(ggalign)
```

Let's begin by creating some example data

```{r setup_dat}
set.seed(123)
small_mat <- matrix(rnorm(81), nrow = 9)
rownames(small_mat) <- paste0("row", seq_len(nrow(small_mat)))
colnames(small_mat) <- paste0("column", seq_len(ncol(small_mat)))
```

`ggalign` provids two layouts to build plots:

  - `layout_heatmap`: heatmap layout
  - `layout_stack`: align plots vertically or horizontally

## `layout_heatmap`
`layout_heatmap` utilizes the grammar of graphics to construct the heatmap and
heatmap annotations. You can also use the alias `ggheatmap`.

```{r}
ggheatmap(small_mat)
```

The data can be a numeric or character vector, a data frame, and any other data
which can be converted into a matrix. Simple vector will be converted into a one
column matrix. They will all converted into the long formated data frame when
drawing. The default mapping will use `aes(.data$.x, .data$.y)`, but can be
controlled using `mapping` argument. The data in the underlying ggplot object
contains following columns:

  - `.row_panel`: the row panel

  - `.column_panel`: the column panel

  - `.row_names` and `.column_names`: the row and column names of the original
    matrix (only applicable when names exist).

  - `.row_index` and `.column_index`: the row and column index of the original
    matrix.

  - `.x` and `.y`: the `x` and `y` coordinates

  - `value`: the actual matrix value.

```{r}
ggheatmap(letters)
```

```{r}
ggheatmap(1:10)
```

```{r}
ggheatmap(small_mat) + geom_point() + scale_fill_viridis_c()
```

You can turn off the default filling by setting `filling = FALSE`
```{r}
ggheatmap(small_mat, filling = FALSE) +
  geom_text(aes(label = sprintf("%d * %d", .row_index, .column_index)))
```

Heatmap annotation works by adding additional information for heatmap
rows/columns. Heatmap annotations can be positioned at the `top`, `left`,
`bottom`, or `right` of the heatmap. This is referred to as the active context
in `ggheatmap`. we can add annotation with `align_*` functions.

## `align_*`

>The `ggalign` package provides a range of `align_*` functions that enable
control over the layout. Specifically, these functions allow for the
manipulation of axis order in the layout, as well as partitioning along a single
axis into multiple panels. Additionally, some of these functions have the
capability to add plots. 

Currently, there are four primary `align_*` functions: `align_group`,
`align_reorder`, `align_kmeans` and `align_dendro`.

### `align_group`

`align_group` just group along a layout axis into different panels.

By default, the `ggheatmap` function does not initialize any active context, so
if you want to add heatmap annotations, you need to use `hmanno` to specify
where the annotation should be added.

```{r align_group_top}
ggheatmap(small_mat) +
  hmanno("t") +
  align_group(sample(letters[1:4], ncol(small_mat), replace = TRUE))
```

By default, the facet strip text will be removed, you can override the behaviour
with `theme(strip.text = element_text())`.
```{r align_group_left}
ggheatmap(small_mat) +
  theme(strip.text = element_text()) +
  hmanno("l") +
  align_group(sample(letters[1:4], nrow(small_mat), replace = TRUE))
```

### `align_reorder`
`align_reorder` reorder heatmap rows/columns based on a summary function.
```{r}
ggheatmap(small_mat) +
  hmanno("l") +
  align_reorder(rowMeans)
```

Some `align_*` functions provide a `data` argument to specify the data. This
argument can accept a matrix, a data frame, or even a simple vector that will be
converted into a one-column matrix. If the `data` argument is set to `NULL`, the
function will use the layout data, as shown in the previous example.
Additionally, the `data` argument can also accept a function (purrr-like lambda
is also okay), which will be applied with the layout data. 

> It is important to note that all `align_*` function consider the rows as the
observations. It means the `NROW` function must return the same number as the
layout parallel axis. So for heatmap column annotation, the heatmap matrix will
be transposed before using (If `data` is a function, it will be applied with the
transposed matrix).

So even for top and bottom annotation, we can also use the `rowMeans` to
calculate the mean value for all columns.

```{r}
ggheatmap(small_mat) +
  hmanno("t") +
  align_reorder(rowMeans)
```

### `align_kmeans`
`align_kmeans` group heatmap rows/columns by kmeans.
```{r}
ggheatmap(small_mat) +
  hmanno("t") +
  align_kmeans(3L)
```

It is important to note that both `align_group` and `align_kmeans` cannot do
sub-groups. It means when there has been any groups exist. You cannot use them

```{r error=TRUE}
ggheatmap(small_mat) +
  hmanno("t") +
  align_group(sample(letters[1:4], ncol(small_mat), replace = TRUE)) +
  align_kmeans(3L)
```

```{r error=TRUE}
ggheatmap(small_mat) +
  hmanno("t") +
  align_kmeans(3L) +
  align_group(sample(letters[1:4], ncol(small_mat), replace = TRUE))
```

### align_dendro
`align_dendro` class the addition of a dendrogram near the heatmap and and it
can also reorder the heatmap. This is primarily useful when working with
heatmap plots.

```{r}
ggheatmap(small_mat) +
  hmanno("t") +
  align_dendro()
```

You can use the `distance` and `method` argument to control the dendrogram
builind proce. 

```{r}
ggheatmap(small_mat) +
  hmanno("t") +
  align_dendro(method = "ward.D2")
```

One useful function of dendrogram is to cut the columns/rows into groups. You
can specify `k` or `h`, which works the same with `cutree`.

```{r}
ggheatmap(small_mat) +
  hmanno("t") +
  align_dendro(k = 3L)
```

In contrast to `align_group`, `align_kmeans`, and `align_reorder`,
`align_dendro` is capable of drawing plot components. Therefore, it has a
default `set_context` value of `TRUE`, which indicates whether to set the active
context to current Align object. we can directly add ggplot elements after
`align_dendro` function. 

```{r}
ggheatmap(small_mat) +
  hmanno("t") +
  align_dendro() +
  geom_point(aes(y = y))
```

`align_dendro` will create a default `node` data for the ggplot. See
`?dendrogram_data` for details. In addition, `edge` data has been added into the
`geom_segment` layer directly which was used to draw the dendrogram tree. One
useful variable in both `node` and edge `data` is `branch` column, which
corresponding to the `cutree` result.

```{r}
ggheatmap(small_mat) +
  hmanno("t") +
  align_dendro(aes(color = branch), k = 3) +
  geom_point(aes(color = branch, y = y))
```

`align_dendro` is also capable of performing clustering between groups. This
means that you can use it even if there are already existing groups present in
the heatmap.

```{r}
column_groups <- sample(letters[1:3], ncol(small_mat), replace = TRUE)
ggheatmap(small_mat) +
  hmanno("t") +
  align_group(column_groups) +
  align_dendro(aes(color = branch))
```

You can reorder the groups by setting `reorder_group = TRUE`.
```{r}
ggheatmap(small_mat) +
  hmanno("t") +
  align_group(column_groups) +
  align_dendro(aes(color = branch), reorder_group = TRUE)
```

You can see the difference.
```{r}
ggheatmap(small_mat) +
  hmanno("t") +
  align_group(column_groups) +
  align_dendro(aes(color = branch), reorder_group = TRUE) +
  hmanno("b") +
  align_dendro(aes(color = branch), reorder_group = FALSE)
```

It is important to understand that when `reorder_group = FALSE` is used, the
reordering of the heatmap occurs within each group. As long as the ordering
within each group remains the same, these two dendrograms can be placed on the
same axis of the heatmap. This is why adding a dendrogram with` reorder_group =
FALSE` behind a dendrogram with `reorder_group = TRUE` is acceptable, as the
second `align_dendro` will follow the group ordering established by the first
one. However, it is not possible to add a dendrogram with `reorder_group = TRUE`
behind a dendrogram with `reorder_group = FALSE` because the second dendrogram
would not be able to conform to the grouping order followed by the first
dendrogram.

```{r error=TRUE}
ggheatmap(small_mat) +
  hmanno("t") +
  align_group(column_groups) +
  align_dendro(aes(color = branch), reorder_group = FALSE) +
  hmanno("b") +
  align_dendro(aes(color = branch), reorder_group = TRUE)
```

>`ggalign` always prevent users from reordering the layout axis twice.

## `align_gg`

`align_gg` is similar to `ggplot` in that it initializes a `ggplot` data and
`mapping`. The data input can be a matrix, a data frame, or a simple vector that
will be converted into a one-column matrix, and can inherit from the heatmap
matrix. But for ggplot usage, matrix (including a simple vector) data is
converted into a long-format data frame, similar to the process utilized in
`ggheatmap`. But note that the long-format data frame does not contain `.row_panel`
or `.column_panel` column, as annotations can only have one facet axis. In the
case where the input data is already a data frame, three additional
columns-`.row_names`, `.row_index`, and `.panel`—are added to the data frame. 

You can also use `ggalign` which is an alias of `align_gg`.

```{r}
ggheatmap(small_mat) +
  scale_fill_viridis_c(guide = "none") +
  hmanno("t") +
  align_dendro(aes(color = branch), k = 3) +
  ggalign(data = rowSums) +
  geom_bar(aes(y = value, fill = .panel), stat = "identity") +
  scale_fill_brewer(palette = "Dark2") +
  hmanno("l") +
  ggalign(aes(x = value), data = rowSums) +
  geom_bar(
    aes(y = .y, fill = factor(.y)),
    stat = "identity",
    orientation = "y"
  ) +
  scale_fill_brewer(palette = "Set1") +
  scale_x_reverse()
```

## Control size
The different ggplot objects were combined using the `patchwork` package.
Internally, the `build_patchwork` function was used to construct and merge all
the `ggplot` objects into a single `patchwork` object. 

`ggheatmap` has `width` and `height` to control the relative (you can also
provide a unit object) width and height of the heatmap body. `hmanno` function
have a `size` argument to control the relative width (left and right annotation)
or height (top and bottom annotation) of the whole annotation. `anno_*` function
has a `size` argument to control the the relative width (left and right annotation)
or height (top and bottom annotation) of the single annotation in the whole
annotation.

```{r}
ggheatmap(small_mat) +
  scale_fill_viridis_c() +
  hmanno("t", size = unit(20, "mm")) +
  ggalign(data = rowSums) +
  geom_bar(aes(y = value, fill = .x), stat = "identity") +
  hmanno("l", size = 0.2) +
  ggalign(aes(x = value), data = rowSums) +
  geom_bar(
    aes(y = .y, fill = factor(.y)),
    stat = "identity",
    orientation = "y"
  ) +
  scale_fill_brewer(palette = "Set1", guide = "none") +
  scale_x_reverse()
```

## restriction
The position scales and facets are partial support.

For position scales, you cannot set `limits`, `breaks`, `labels`, and they are
just ignored. `limits` cannot be touched, and the internal will always reset it
as the default. This is required to match the heatmap and annotation limits. But
you can set `labels` or `breaks` in `ggheatmap()` function. See `xlabels/ylabels`
and `xlabels_nudge/ylabels_nudge`. All of these arguments should be provided in
the original order of the raw matrix.
```{r}
ggheatmap(small_mat) + scale_x_continuous(limits = c(0, 0))
```

```{r}
ggheatmap(small_mat) + scale_x_continuous(labels = "a")
```

```{r}
ggheatmap(small_mat, xlabels = rep_len("AA", ncol(small_mat)))
```

When working with facets, manual configuration of panels using the `facet_*`
function is not possible since the internal will use `facet_grid` to set the
row/column groups controlled by heatmap annotation. However, you can also
provide `facet_grid` or `facet_null` (if no panels) to control the non-layout
arguments, like `labeller`.

```{r}
ggheatmap(small_mat) +
  facet_grid(labeller = labeller(.panel = function(x) letters[as.integer(x)])) +
  hmanno("t") +
  align_dendro(aes(color = branch), k = 3) +
  geom_point(aes(y = y))
```

## Session information
```{r}
sessionInfo()
```
