---
title: "Introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# ggheat

This package is a `ggplot2` extension for heatmap. It utilizes the grammar of
graphics to construct the heatmap and heatmap annotations.

```{r setup}
library(ggheat)
```

Let's begin by creating some example data

```{r setup_dat}
set.seed(123)
small_mat <- matrix(rnorm(81), nrow = 9)
rownames(small_mat) <- paste0("row", seq_len(nrow(small_mat)))
colnames(small_mat) <- paste0("column", seq_len(ncol(small_mat)))
```

## `ggheat`
The core function in the `ggheat` package is `ggheat`, which sets the default data
and mapping for the heatmap. Creating a simple heatmap is effortless with the
`ggheat` function 

```{r}
ggheatmap(small_mat)
```

The data can be a numeric or character vector, a data frame, and any other data
which can be converted into a matrix. Simple vector will be converted into a one
column matrix. They will all converted into the long formated data frame when
drawing. The default mapping will use `aes(.data$.x, .data$.y)`, but can be
controlled using `mapping` argument. The data in the underlying ggplot object
contains following columns:

  - `.row_panel`: the row panel

  - `.column_panel`: the column panel

  - `.row_names` and `.column_names`: the row and column names of the original
    matrix (only applicable when names exist).

  - `.row_index` and `.column_index`: the row and column index of the original
    matrix.

  - `.x` and `.y`: the `x` and `y` coordinates

  - `value`: the actual matrix value.

```{r}
ggheatmap(letters)
```

```{r}
ggheatmap(1:10)
```

You can then add all other ggplot2 elements like `geoms`, `scales` and `facets`.

```{r}
ggheatmap(small_mat) + geom_point() + scale_fill_viridis_c()
```

You can turn off the default filling by setting `filling = FALSE`
```{r}
ggheatmap(small_mat, filling = FALSE) +
  geom_text(aes(label = sprintf("%d * %d", .row_index, .column_index)))
```

Heatmap annotation works by adding additional information for heatmap
rows/columns. Heatmap annotations can be positioned at the `top`, `left`,
`bottom`, or `right` of the heatmap. This is referred to as the active context
in `ggheat`.

>`ggheat` offers two primary annotations: `htanno` and `gganno`. 

## `htanno`

`htanno` is a special annotation that can interact with the main heatmap. It
particularly allows for controlling the order of the main heatmap or dividing
the heatmap into sections based on rows/columns. Some of them can also adding
plot.

Currently, there are four sub-classes of `htanno`: `align_group`,
`align_reorder`, `align_kmeans` and `align_dendro`.

### `align_group`

`align_group` just group heatmap rows/columns into different panels.

By default, the `ggheat` function does not initialize any active context, so if
you want to add annotations, you need to specify the `position` argument. We can
omit it if there is any active context in the heatmap (we'll introduce how to
set the active context after for details). 
```{r align_group_top}
ggheatmap(small_mat) +
  hmanno("t") +
  align_group(sample(letters[1:4], ncol(small_mat), replace = TRUE))
```

```{r align_group_left}
ggheatmap(small_mat) +
  hmanno("l") +
  align_group(sample(letters[1:4], nrow(small_mat), replace = TRUE))
```

### `align_reorder`
`align_reorder` reorder heatmap rows/columns based on a summary function.
```{r}
ggheatmap(small_mat) +
  hmanno("l") +
  align_reorder(rowMeans)
```

Some annotation function includes a `data` argument to specify the annotation
data. This argument can accept a matrix, a data frame, or even a simple vector
that will be converted into a one-column matrix. If the `data` argument is set
to `NULL`, the function will use the heatmap matrix, as shown in the previous
example. Additionally, the `data` argument can also accept a function (purrr-like
lambda is also okay), which will be applied with the heatmap matrix. 

> It is important to note that all annotations consider the rows as the
> observations. It means the `NROW` function must return the same number as the
> heatmap parallel axis. So for column annotation, the heatmap will be transposed
> before using (If `data` is a function, it will be applied with the transposed
> matrix).

So even for top and bottom annotation, we can also use the `rowMeans` to calculate
the mean value for all columns.

```{r}
ggheatmap(small_mat) +
  hmanno("t") +
  align_reorder(rowMeans)
```

### `align_kmeans`
`align_kmeans` group heatmap rows/columns by kmeans.
```{r}
ggheatmap(small_mat) +
  hmanno("t") +
  align_kmeans(3L)
```

It is important to note that both `align_group` and `align_kmeans` cannot do
sub-groups. It means when there has been any groups exist. You cannot use them

```{r error=TRUE}
ggheatmap(small_mat) +
  hmanno("t") +
  align_group(sample(letters[1:4], ncol(small_mat), replace = TRUE)) +
  align_kmeans(3L)
```

```{r error=TRUE}
ggheatmap(small_mat) +
  hmanno("t") +
  align_kmeans(3L) +
  align_group(sample(letters[1:4], ncol(small_mat), replace = TRUE))
```

You may have noticed that the `align_group` function does not explicitly set
the `position` argument after `align_kmeans`. When adding an annotation, we can
set the active context using the `set_context` argument. This argument requires
a boolean value of length `2`. It will be recycled for a single value.

 - The first value determines whether the active context is set to
   the annotation `position`.
 - The second value will determine whether the active context of the annotation
   list in this position should be set as current annotation.

In the case of `align_kmeans`, the default `set_context` value is `c(TRUE,
FALSE)`, implying that the active context is already set by `align_kmeans`. 

### align_dendro
`align_dendro` class the addition of a dendrogram near the heatmap and and it
can also reorder the heatmap. This is primarily useful when working with
heatmap plots.

```{r}
ggheatmap(small_mat) +
  hmanno("t") +
  align_dendro()
```

You can use the `distance` and `method` argument to control the dendrogram
builind proce. 

```{r}
ggheatmap(small_mat) +
  hmanno("t") +
  align_dendro(method = "ward.D2")
```

One useful function of dendrogram is to cut the columns/rows into groups. You
can specify `k` or `h`, which works the same with `cutree`.

```{r}
ggheatmap(small_mat) +
  hmanno("t") +
  align_dendro(k = 3L)
```

In contrast to `align_group`, `align_kmeans`, and `align_reorder`,
`align_dendro` is capable of drawing plot components. Therefore, it has a
default `set_context` value of `c(TRUE, TRUE)`, we can directly add ggplot
elements after `align_dendro` function. 

```{r}
ggheatmap(small_mat) +
  hmanno("t") +
  align_dendro() +
  geom_point(aes(y = y))
```

`align_dendro` will create a default `node` data for the ggplot. See
`?dendrogram_data` for details. In addition, `edge` data has been added into the
`geom_segment` layer directly which was used to draw the dendrogram tree. One
useful variable in both `node` and edge `data` is `branch` column, which
corresponding to the `cutree` result.

```{r}
ggheatmap(small_mat) +
  hmanno("t") +
  align_dendro(aes(color = branch), k = 3) +
  geom_point(aes(color = branch, y = y))
```

`align_dendro` is also capable of performing clustering between groups. This
means that you can use it even if there are already existing groups present in
the heatmap.

```{r}
column_groups <- sample(letters[1:3], ncol(small_mat), replace = TRUE)
ggheatmap(small_mat) +
  hmanno("t") +
  align_group(column_groups) +
  align_dendro(aes(color = branch))
```

You can reorder the groups by setting `reorder_group = TRUE`.
```{r}
ggheatmap(small_mat) +
  hmanno("t") +
  align_group(column_groups) +
  align_dendro(aes(color = branch), reorder_group = TRUE)
```

You can see the difference.
```{r}
ggheatmap(small_mat) +
  hmanno("t") +
  align_group(column_groups) +
  align_dendro(aes(color = branch), reorder_group = TRUE) +
  hmanno("b") +
  align_dendro(aes(color = branch), reorder_group = FALSE)
```

It is important to understand that when `reorder_group = FALSE` is used, the
reordering of the heatmap occurs within each group. As long as the ordering
within each group remains the same, these two dendrograms can be placed on the
same axis of the heatmap. This is why adding a dendrogram with` reorder_group =
FALSE` behind a dendrogram with `reorder_group = TRUE` is acceptable, as the
second `align_dendro` will follow the group ordering established by the first
one. However, it is not possible to add a dendrogram with `reorder_group = TRUE`
behind a dendrogram with `reorder_group = FALSE` because the second dendrogram
would not be able to conform to the grouping order followed by the first
dendrogram.

```{r error=TRUE}
ggheatmap(small_mat) +
  hmanno("t") +
  align_group(column_groups) +
  align_dendro(aes(color = branch), reorder_group = FALSE) +
  hmanno("b") +
  align_dendro(aes(color = branch), reorder_group = TRUE)
```

>`ggheat` always prevent user from reordering the heatmap twice.

## `gganno`

`gganno` is similar to `ggheat` and `ggplot` in that it initializes a `ggplot`
data and `mapping`. The data input can be a matrix, a data frame, or a simple
vector that will be converted into a one-column matrix, and can inherit from the
heatmap matrix. But for ggplot usage, matrix (including a simple vector) data is
converted into a long-format data frame, similar to the process utilized in
`ggheat`. But note that the long-format data frame does not contain `.row_panel`
or `.column_panel` column, as annotations can only have one facet axis. In the
case where the input data is already a data frame, three additional
columns-`.row_names`, `.row_index`, and `.panel`—are added to the data frame. 

Let's add more annotations to different position. Since we want to add
annotation into another position, we must specify the `position` argument or we
can add an `active` object (we also provide `activate` and `deactivate`
for function call foramt).

```{r}
ggheatmap(small_mat) +
  scale_fill_viridis_c(guide = "none") +
  hmanno("t") +
  align_dendro(aes(color = branch), k = 3) +
  ggalign(data = rowSums) +
  geom_bar(aes(y = value, fill = .panel), stat = "identity") +
  scale_fill_brewer(palette = "Dark2") +
  hmanno("l") +
  ggalign(aes(x = value), data = rowSums) +
  geom_bar(
    aes(y = .y, fill = factor(.y)),
    stat = "identity",
    orientation = "y"
  ) +
  scale_fill_brewer(palette = "Set1") +
  scale_x_reverse()
```

## Control size
The different ggplot objects were combined using the `patchwork` package.
Internally, the `ggheat_build` function was used to construct and merge all the
`ggplot` objects into a single `patchwork` object. 

`ggheat` has `width` and `height` to control the relative (you can also provide
a unit object) width and height. All annotation function have a `size` argument
to control the relative width (left and right annotation) or height (top and
bottom annotation).

```{r}
ggheatmap(small_mat) +
  scale_fill_viridis_c() +
  hmanno("t", size = unit(20, "mm")) +
  ggalign(data = rowSums) +
  geom_bar(aes(y = value, fill = .x), stat = "identity") +
  hmanno("l", size = 0.2) +
  ggalign(aes(x = value), data = rowSums) +
  geom_bar(
    aes(y = .y, fill = factor(.y)),
    stat = "identity",
    orientation = "y"
  ) +
  scale_fill_brewer(palette = "Set1", guide = "none") +
  scale_x_reverse()
```

## restriction
The position scales and facets are partial support.

For position scales, you cannot set `limits`, `breaks`, `labels`, and they are
just ignored. `limits` cannot be touched, and the internal will always reset it
as the default. This is required to match the heatmap and annotation limits. But
you can set `labels` or `breaks` in `ggheatmap()` function. See `xlabels/ylabels`
and `xlabels_nudge/ylabels_nudge`. All of these arguments should be provided in
the original order of the raw matrix. Even if we use heatmap annotation to
reorder the heatmap rows/columns. 
```{r}
ggheatmap(small_mat) + scale_x_continuous(limits = c(0, 0))
```

```{r}
ggheatmap(small_mat) + scale_x_continuous(labels = "a")
```

```{r}
ggheatmap(small_mat, xlabels = rep_len("AA", ncol(small_mat)))
```

When working with facets, manual configuration of panels using the `facet_*`
function is not possible since the internal will use `facet_grid` to set the
row/column groups controlled by heatmap annotation. However, you can also
provide `facet_grid` or `facet_null` (if no panels) to control the non-layout
arguments, like `labeller`.

```{r}
ggheatmap(small_mat) +
  facet_grid(labeller = labeller(.panel = function(x) letters[as.integer(x)])) +
  hmanno("t") +
  align_dendro(aes(color = branch), k = 3) +
  geom_point(aes(y = y))
```

## Session information
```{r}
sessionInfo()
```
