---
title: "Stack Layout"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Stack Layout}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`layout_stack` put plots horizontally or vertically. You can also use the alias
`ggstack`.

```{r setup}
library(ggalign)
```

 You can also specify which plot context of the annotation to use with the
`what` argument in `hmanno()`. However, this is usually not necessary as each
`align_*` function has a `set_context` argument that controls whether it sets
the active context to the `Align` object when added to the stack layout:

- For `align_*` functions that draw something (e.g., `align_dendro()`,
  `ggalign()`), `set_context` is set to `TRUE` by default. 
- For `align_*` functions that do not draw anything (e.g., `align_group()`,
  `align_kmeans()`), `set_context` is set to `FALSE` by default.
```{r}

```

## Session information
```{r}
sessionInfo()
```
