---
title: "Plot options"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plot options}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

Plot options control the actions of plots within the layout. These options can
be applied either globally to all plots in the layout or individually to
specific plots.

 - To apply a plot option to a single plot, use the `+` operator.
 - To set a plot option at the layout level (affecting all plots), use the `-`
 operator. 

>Plot options inherit properties from parent layout options hierarchically.

The package currently provides three plot options, each prefixed with `plot_`:

  - `plot_theme`: Sets the default theme for the plot.
  - `plot_data`: Transforms the plot data. Many functions in this package
  require a specific data format to align observations, `plot_data()` helps
  reformat data frames as needed.
  - `plot_align`: Defines alignment specifications for plots within the layout.

```{r setup}
library(ggalign)
```

```{r setup_data}
set.seed(123)
small_mat <- matrix(rnorm(81), nrow = 9)
rownames(small_mat) <- paste0("row", seq_len(nrow(small_mat)))
colnames(small_mat) <- paste0("column", seq_len(ncol(small_mat)))
```

# `plot_theme`
The `plot_theme()` function extends `theme()` to set a default theme for plots,
allowing you to input `theme()` elements directly or modify the theme elements.

To set a plot option for a single plot, simply use the `+` operator:
```{r}
ggheatmap(small_mat) +
    # change the default theme of the heatmap body
    plot_theme(plot.background = element_rect(fill = "red"))
```

Using a `theme()` Object in `plot_theme()`.
```{r}
ggheatmap(small_mat, filling = FALSE) +
    geom_tile(aes(fill = value), width = 0.9, height = 0.9) +
    # change the default theme of the heatmap body
    plot_theme(theme_bw(), plot.background = element_rect(fill = "red"))
```

Note that `plot_theme()` serves as the default theme and will always be
overridden by any `theme()` settings applied directly to the plot. The default
theme (`plot_theme()`) is applied first, followed by any specific `theme()`
settings, even if `theme()` is added before `plot_theme()`.
```{r}
ggheatmap(small_mat) +
    # change the plot theme of the heatmap body
    theme(plot.background = element_rect(fill = "blue")) +
    # change the default theme of the heatmap body
    plot_theme(plot.background = element_rect(fill = "red"))
```

By using the `-` operator with plot options, we apply the option directly to the
active layout.
```{r}
ggheatmap(small_mat) +
    # Change the active layout to the top annotation
    anno_top() +
    # add a dendrogram to the top annotation
    align_dendro() +
    # add a bar plot to the top annotation
    ggalign(aes(.names, value, fill = factor(.names)), data = rowSums) +
    geom_bar(stat = "identity") -
    # Change the default theme of the top annotation
    # All plots in the top annotation will inherit this default theme
    plot_theme(plot.background = element_rect(fill = "red"))
```

Unlike individual ggplot2 elements added to each plot, layout-level options set
by `-` operator are inherited by all plots in the layout when rendered. Any
plot-specific options will override these layout-level options, regardless of
the order in which they are added.

```{r}
ggheatmap(small_mat) +
    # Change the active layout to the top annotation
    anno_top() +
    # add a dendrogram to the top annotation
    align_dendro() +
    # change the plot_theme for the dendrogram plot
    plot_theme(plot.background = element_rect(fill = "blue")) +
    # add a bar plot to the top annotation
    ggalign(aes(.names, value, fill = factor(.names)), data = rowSums) +
    geom_bar(stat = "identity") -
    # Change the default theme of the top annotation
    # All plots in the top annotation will inherit this default theme
    # But the plot-specific options will override these
    plot_theme(plot.background = element_rect(fill = "red"))
```

# `plot_data`
`align_gg()`/`ggalign()` requires the specific data format for its operations.
If you need to transform or filter data for individual `geoms`, you can use the
`data` argument within each `geom`. However, if you have multiple `geoms` and
want a consistent transformation applied across all, you can utilize the
`plot_data()` function. This allows you to transform the default data for the
entire plot.

The `plot_data()` accepts a function that takes a data frame as input and
returns a modified data frame. By default, `plot_data()` will attempt to inherit
from the parent layout if the data is inherited from it. However, there is one
exception: `align_dendro()` will not inherit `plot_data()` transformations by
default.

```{r}
set.seed(1234L)
ggheatmap(small_mat) +
    anno_top() +
    align_kmeans(3L) +
    # we add a bar plot
    ggalign() +
    # we subest the plot data
    plot_data(~ subset(.x, .panel == 1L)) +
    geom_bar(aes(y = value, fill = .row_names), stat = "identity")
```

# `plot_align`
The `plot_align()` function defines the align Specifications for plots.

## guides
By default, `ggheatmap()` will collect all guide legends on the side from which
they originate. 

```{r fig.dim = c(12, 12)}
heatmap_collect_all_guides <- ggheatmap(small_mat, width = 2, height = 2L) +
    scale_fill_viridis_c(name = "I'm from heatmap body") +
    theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
    anno_top() +
    align_dendro(aes(color = branch), k = 3L) +
    scale_color_brewer(
        name = "I'm from top annotation", palette = "Dark2",
        guide = guide_legend(position = "right")
    ) +
    anno_left() +
    align_dendro(aes(color = branch), k = 3L) +
    scale_color_brewer(
        name = "I'm from left annotation", palette = "Dark2",
        guide = guide_legend(position = "top", direction = "vertical")
    ) +
    anno_bottom() +
    align_dendro(aes(color = branch), k = 3L) +
    scale_color_brewer(
        name = "I'm from bottom annotation", palette = "Dark2",
        guide = guide_legend(position = "left")
    ) +
    anno_right() +
    align_dendro(aes(color = branch), k = 3L) +
    scale_color_brewer(
        name = "I'm from right annotation", palette = "Dark2",
        guide = guide_legend(position = "bottom", direction = "vertical")
    ) &
    theme(plot.margin = margin())
heatmap_collect_all_guides
```

Use the `guides` argument control which side of guide legends should be
gathered. In the following example, we'll collect the guide legends only on the
`top` and `bottom` sides: 
```{r fig.dim = c(12, 12)}
heatmap_collect_all_guides +
    # reset the active context to the heatmap layout
    quad_active() -
    # we set global `guides` argument for the heatmap layout
    # we only collect guides in the top and bottom side
    plot_align(guides = "tb")
```

You can also apply the `plot_align()` function directly to specific plots:
```{r fig.dim = c(12, 12)}
heatmap_collect_all_guides +
    # reset the active context to the heatmap layout
    quad_active() -
    # we set global `guides` argument for the heatmap layout
    # we only collect guides in the top and bottom side
    plot_align(guides = "tb") +
    # `+` apply it to the current active plot
    # for the heatmap body, we collect guide in the right side
    plot_align(guides = "r")
```

The guide legends within the annotation stack are first collected by the
`stack_layout()` itself and then passed to the overall heatmap layout for
further integration. By default, it inherits the `guides` arguments from the
heatmap layout. See following example: 
```{r fig.dim = c(12, 12)}
heatmap_collect_all_guides +
    # reset the active context to the heatmap layout
    quad_active() -
    # we set global `guides` argument for the heatmap layout
    # we only collect guides in the top and bottom side
    plot_align(guides = "tb") +
    # we ensure the active context is in the bottom annotation
    # By default, it inherits "guides" argument from the heamtap layout, which
    # means it'll collect "guides" in the top and bottom side
    anno_bottom() +
    # for the dendrogram in the bottom annotation, we collect guide in the left side
    plot_align(guides = "l")
```

Here, the guide legend is collected by the bottom annotation but will not be
collected by the heatmap layout since the heatmap layout only gathers guides
from the `top` and `bottom`. In this way, the guide legends of the annotation
stack will be put around the annotation stack layout.

To override this guide collection behavior for the heatmap annotation, you can
use the `free_guides` argument of the `quad_anno()`/`anno_*()` function. This
differs from the `guides` argument in `plot_align()`, which controls the
behavior for the plots in the layout. The `free_guides` argument specifies which
guide legends from the annotation stack layout should be collected by the
heatmap layout. 
```{r fig.dim = c(12, 12)}
heatmap_collect_all_guides +
    # reset the active context to the heatmap layout
    quad_active() -
    # we set global `guides` argument for the heatmap layout
    # we only collect guides in the top and bottom side
    plot_align(guides = "tb") +
    # we also collect guides in the left side for the bottom annotation stack
    # layout in the heatmap layout
    anno_bottom(free_guides = "l") +
    # for the dendrogram in the bottom annotation, we collect guide in the left side
    plot_align(guides = "l")
```

>Note: The heatmap layout will only collect guide legends from the annotation
stack if the stack layout collects its own guides first.

## free_spaces
By default, `ggheatmap()` will align all elements of the plot, which can
sometimes lead to unwanted spacing. Consider the following example:

```{r}
ggheatmap(small_mat) +
    scale_fill_viridis_c() +
    anno_top(size = unit(30, "mm")) +
    align_dendro() +
    scale_y_continuous(
        expand = expansion(),
        labels = ~ paste("very very long labels", .x)
    ) +
    anno_left(unit(20, "mm")) +
    align_dendro()
```

In this case, the left annotation stack is positioned far from the heatmap body
due to the wide axis labels in the top annotation stack. This occurs because the
top annotation stack is aligned with the heatmap body. To fix this, you can
remove the left borders around the panel of the top annotation stack by setting
`free_spaces = "l"`.

```{r}
ggheatmap(small_mat) +
    scale_fill_viridis_c() +
    anno_top(size = unit(30, "mm")) -
    plot_align(free_spaces = "l") +
    align_dendro() +
    scale_y_continuous(
        expand = expansion(),
        labels = ~ paste("very long labels", .x)
    ) +
    anno_left(unit(20, "mm")) +
    align_dendro()
```

One useful way to utilize `free_spaces` is to position the guide legends next to
the annotations. (Note the guide legend from the bottom annotation): 
```{r fig.dim = c(12, 12)}
heatmap_collect_all_guides +
    # we only collect guides in the top and bottom side
    quad_active() -
    plot_align(guides = "tb") +
    # 1. in the bottom annotation stack layout, we collect the legends in the
    #    left side
    # 2. we remove the spaces of the left border in the annotation
    anno_bottom() -
    plot_align(guides = "l", free_spaces = "l")
```

In `ggheatmap()`/`quad_layout()`, the behavior of the `free_spaces` and
`free_labs` arguments differs from `guides` arguments in `plot_align()` when
inheriting from the parent layout:
  - For `top` and `bottom` annotations, it inherits from the left ("l") and
  right ("r") axes.
  - For `left` and `right` annotations, it inherits from the top ("t") and
  bottom ("b") axes.

## free_labs
By default, we won't align the axis titles.
```{r}
ggheatmap(small_mat) +
    scale_fill_viridis_c() +
    ylab("Heatmap title") +
    anno_top(size = unit(30, "mm")) +
    align_dendro() +
    ylab("Annotation title")
```

To align axis titles, you can set `free_labs = NULL`. Alternatively, A single
string containing one or more of axis positions ("t", "l", "b", "r") to indicate
which axis titles should be free from alignment.

```{r}
ggheatmap(small_mat) -
    plot_align(free_labs = NULL) +
    scale_fill_viridis_c() +
    ylab("Heatmap title") +
    anno_top(size = unit(30, "mm")) +
    align_dendro() +
    ylab("Annotation title")
```

## Session information
```{r}
sessionInfo()
```
