---
title: "Plot Options"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plot Options}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

Plot options control the actions of plots within the layout. These options can
be applied either globally to all plots in the layout or individually to
specific plots.

 - To apply a plot option to a single plot, use the `+` operator.
 - To set a plot option at the layout level (affecting all plots), use the `-`
 operator. 

>Plot options inherit properties from parent layout options hierarchically.

The package currently provides three plot options, each prefixed with `plot_`:

  - `plot_theme`: Sets the default theme for the plot.
  - `plot_data`: Transforms the plot data. Many functions in this package
  require a specific data format to align observations, `plot_data()` helps
  reformat data frames as needed.
  - `plot_align`: Defines alignment specifications for plots within the layout.

```{r setup}
library(ggalign)
```

```{r setup_data}
set.seed(123)
small_mat <- matrix(rnorm(81), nrow = 9)
rownames(small_mat) <- paste0("row", seq_len(nrow(small_mat)))
colnames(small_mat) <- paste0("column", seq_len(ncol(small_mat)))
```

# `plot_theme`
The `plot_theme()` function extends `theme()` to set a default theme for plots,
allowing you to input `theme()` elements directly or modify the theme elements.

To set a plot option for a single plot, simply use the `+` operator:
```{r}
ggheatmap(small_mat) +
    # change the default theme of the heatmap body
    plot_theme(plot.background = element_rect(fill = "red"))
```

You can use a `theme()` object directly in `plot_theme()`.
```{r}
ggheatmap(small_mat, filling = FALSE) +
    # we reduce the heatmap body cell width and height
    # in this way, we can see the background of `theme_bw()` in the panel.
    geom_tile(aes(fill = value), width = 0.9, height = 0.9) +
    # change the default theme of the heatmap body
    plot_theme(theme_bw(), plot.background = element_rect(fill = "red"))
```

Note that `plot_theme()` serves as the default theme and will always be
overridden by any `theme()` settings applied directly to the plot. The default
theme (`plot_theme()`) is applied first, followed by any specific `theme()`
settings, even if `theme()` is added before `plot_theme()`.
```{r}
ggheatmap(small_mat) +
    # change the plot theme of the heatmap body
    theme(plot.background = element_rect(fill = "blue")) +
    # change the default theme of the heatmap body
    plot_theme(plot.background = element_rect(fill = "red"))
```

By using the `-` operator with plot options, we apply the option directly to the
active layout.
```{r}
ggheatmap(small_mat) +
    # Change the active layout to the top annotation
    anno_top() +
    # add a dendrogram to the top annotation
    align_dendro() +
    # add a bar plot to the top annotation
    ggalign(aes(.names, value, fill = factor(.names)), data = rowSums) +
    geom_bar(stat = "identity") -
    # Change the default theme of the top annotation
    # All plots in the top annotation will inherit this default theme
    plot_theme(plot.background = element_rect(fill = "red"))
```

Unlike individual ggplot2 elements, which will be added directly to each plot by
`-` operator, **layout-level options** set by `-` operator are **inherited** by
all plots in the layout when `rendered`. Any plot-specific options will override
these layout-level options, regardless of the order in which they are added.

```{r}
ggheatmap(small_mat) +
    # Change the active layout to the top annotation
    anno_top() +
    # add a dendrogram to the top annotation
    align_dendro() +
    # change the plot_theme for the dendrogram plot
    plot_theme(plot.background = element_rect(fill = "blue")) +
    # add a bar plot to the top annotation
    ggalign(aes(.names, value, fill = factor(.names)), data = rowSums) +
    geom_bar(stat = "identity") -
    # Change the default theme of the top annotation
    # All plots in the top annotation will inherit this default theme
    # But the plot-specific options will override these
    plot_theme(plot.background = element_rect(fill = "red"))
```

# `plot_data`
`align_gg()`/`ggalign()` requires the specific data format for its operations.
If you need to transform or filter data for individual `geoms`, you can use the
`data` argument within each `geom`. However, if you have multiple `geoms` and
want a consistent transformation applied across all, you can utilize the
`plot_data()` function. This allows you to transform the default data for the
entire plot.

The `plot_data()` accepts a function that takes a data frame as input and
returns a modified data frame. By default, `plot_data()` will attempt to inherit
from the parent layout if the data is inherited from it. However, there is one
exception: `align_dendro()` will not inherit `plot_data()` transformations by
default.

```{r}
set.seed(1234L)
ggheatmap(small_mat) +
    anno_top() +
    align_kmeans(3L) +
    # we add a bar plot
    ggalign() +
    # we subest the plot data
    plot_data(~ subset(.x, .panel == 1L)) +
    geom_bar(aes(y = value, fill = .row_names), stat = "identity")
```

# `plot_align`
The `plot_align()` function defines the align Specifications for plots.

## guides
By default, `ggheatmap()` will collect all guide legends on the side from which
they originate. 

```{r fig.dim = c(12, 12)}
heatmap_collect_all_guides <- ggheatmap(small_mat, width = 2, height = 2L) +
    # we set the legend to the left side
    scale_fill_viridis_c(
        name = "I'm from heatmap body",
        guide = guide_legend(position = "left")
    ) +
    theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
    # we add a top annotation
    anno_top() +
    # in the top annotation, we add a dendrogram
    align_dendro(aes(color = branch), k = 3L) +
    # we set the legends of the dendrogram to the left side
    scale_color_brewer(
        name = "I'm from top annotation", palette = "Dark2",
        guide = guide_legend(position = "left")
    ) +
    # we add a left annotation
    anno_left() +
    align_dendro(aes(color = branch), k = 3L) +
    # we set the legends of the dendrogram to the top side
    scale_color_brewer(
        name = "I'm from left annotation", palette = "Dark2",
        guide = guide_legend(position = "top", direction = "vertical")
    ) &
    # we remove all margins for all plots
    theme(plot.margin = margin())
heatmap_collect_all_guides
```

The guides argument controls which side of guide legends should be gathered. In
the following example, we'll collect the guide legends only on the top (`t`)
sides:
```{r fig.dim = c(12, 12)}
heatmap_collect_all_guides -
    # we set global `guides` argument for `the heatmap layout`
    # we only collect guides in the top side
    with_quad(plot_align(guides = "t"), NULL)
```

You can also apply the `plot_align()` function directly to specific plots:
```{r fig.dim = c(12, 12)}
heatmap_collect_all_guides +
    # reset the active context to the heatmap layout
    quad_active() -
    # we set global `guides` argument for the heatmap layout
    # we only collect guides in the top side
    plot_align(guides = "t") +
    # `+` apply it to the current active plot
    # for the heatmap body, we collect guide in the left side
    plot_align(guides = "l")
```

Note: The legend on the left side of the heatmap body is collected and
positioned on the left side at the layout level.

If you're annoyed by the large space between the left annotation and the heatmap
body, don't worry! This issue can be addressed using the `free_spaces` feature,
which is covered in its dedicated section.

Let's dive deeper into the guide collection process.

In the last example, we set the `guides` argument for the heatmap body. But what
happens when we set the `guides` for the dendrogram in the top annotation? 
```{r}
heatmap_collect_all_guides -
    # we set global `guides` argument for `the heatmap layout`
    # we only collect guides in the top side in the heatmap layout
    with_quad(plot_align(guides = "t"), NULL) +
    anno_top() +
    # `+` apply it to the current active plot
    # for the dendrogram in the top annotation, we collect guide in the left side
    plot_align(guides = "l")
```

Nothing seems to have changed, right? This is because guide legends within
annotation stacks are first collected by the annotation `stack_layout()` and
only then passed to the top-level layout for further integration. 

By default, the annotation stack inherits the `guides` arguments from the
heatmap layout, followed by the inherited of individual plot in the annotation.
So `guides` argument set at top-level (heatmap layout) will affect all guide
collection behaviour.

In this example:

 - The legend on the left side of the dendrogram in the top annotation is
 collected first at the annotation level. 
 - Since the heatmap layout is not set to collect legends from the left side, it
 remains at the left side within the annotation stack. 
 - For this specific case, the top annotation contains only one plot, so its
 annotation-level placement is identical to plot-level placement.  

To override this, you can use the `free_guides` argument of the
`quad_anno()`/`anno_*()` function. This differs from the `guides` argument in
`plot_align()`, which controls the behavior for the plots in the layout. The
`free_guides` argument specifies which guide legends from at the annotation stack
layout level should be collected by the heatmap layout. 
```{r fig.dim = c(12, 12)}
heatmap_collect_all_guides -
    # we set global `guides` argument for `the heatmap layout`
    # we only collect guides in the top side in the heatmap layout
    with_quad(plot_align(guides = "t"), NULL) +
    # we also collect guides in the left side for the top annotation stack
    # in the heatmap layout
    anno_top(free_guides = "l") +
    # `+` apply it to the current active plot
    # for the dendrogram in the top annotation, we collect guide in the left side
    plot_align(guides = "l")
```

>Note: The heatmap layout will only collect guide legends from the annotation
stack if the stack layout collects its own guides first.

## free_spaces
By default, `ggheatmap()` will align all elements of the plot, which can
sometimes lead to unwanted spacing. Consider the following example:

```{r}
ggheatmap(small_mat) +
    # add top annotation
    anno_top(size = unit(30, "mm")) +
    # add a dendrogram to the top annotation
    align_dendro() +
    # here, we use long labels for visual example
    scale_y_continuous(
        expand = expansion(),
        labels = ~ paste("very very long labels", .x)
    ) +
    # add left annotation
    anno_left(unit(20, "mm")) +
    # add a dendrogram to the left annotation
    align_dendro()
```

In this case, the left annotation stack is positioned far from the heatmap body
due to the wide axis labels in the top annotation stack. This occurs because the
top annotation stack is aligned with the heatmap body. To fix this, you can
remove the left borders around the panel of the top annotation stack by setting
`free_spaces = "l"`.

```{r}
ggheatmap(small_mat) +
    # add top annotation
    anno_top(size = unit(30, "mm")) -
    # we remove the spaces of the left borders in the top annotation
    plot_align(free_spaces = "l") +
    # add a dendrogram to the top annotation
    align_dendro() +
    # here, we use long labels for visual example
    scale_y_continuous(
        expand = expansion(),
        labels = ~ paste("very very long labels", .x)
    ) +
    # add left annotation
    anno_left(unit(20, "mm")) +
    # add a dendrogram to the left annotation
    align_dendro()
```

One useful way to utilize `free_spaces` is to position the guide legends next to
the annotations. (Note the guide legend from the bottom annotation): 

```{r fig.dim = c(12, 12)}
heatmap_collect_all_guides +
    # reset the active context to the heatmap layout
    quad_active() -
    # we set global `guides` argument for the heatmap layout
    # we only collect guides in the top side
    plot_align(guides = "t") +
    # `+` apply it to the current active plot
    # for the heatmap body, we collect guide in the left side
    plot_align(guides = "l") -
    with_quad(plot_align(free_spaces = "l"), "t")
```

Usually you want to apply `free_spaces` with the whole layout, instead of
individual plots.

In `ggheatmap()`/`quad_layout()`, the behavior of the `free_spaces` and
`free_labs` arguments differs from `guides` arguments in `plot_align()` when
inheriting from the parent layout:

  - For `top` and `bottom` annotations, it inherits from the left ("l") and
  right ("r") axes.
  - For `left` and `right` annotations, it inherits from the top ("t") and
  bottom ("b") axes.

## free_labs
By default, we won't align the axis titles.
```{r}
ggheatmap(small_mat) +
    scale_fill_viridis_c() +
    ylab("Heatmap title") +
    anno_top(size = unit(30, "mm")) +
    align_dendro() +
    ylab("Annotation title")
```

To align axis titles, you can set `free_labs = NULL`. Alternatively, A single
string containing one or more of axis positions ("t", "l", "b", "r") to indicate
which axis titles should be free from alignment.

```{r}
ggheatmap(small_mat) -
    plot_align(free_labs = NULL) +
    scale_fill_viridis_c() +
    ylab("Heatmap title") +
    anno_top(size = unit(30, "mm")) +
    align_dendro() +
    ylab("Annotation title")
```

## Session information
```{r}
sessionInfo()
```
