---
title: "Plot options"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plot options}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

Waiting for complete ...

Currently, the package provides three actions (each prefixed with `plot_`):
  - `plot_data`: Transforms the plot data. Many functions in this package
  require a specific data format to align observations, `plot_data()` helps
  reformat data frames as needed. 
  - `plot_align`: Defines alignment specifications for plots within the layout.
  - `plot_theme`: Sets the default theme for the plot.

```{r setup}
library(ggalign)
```

```{r setup_data}
set.seed(123)
small_mat <- matrix(rnorm(81), nrow = 9)
rownames(small_mat) <- paste0("row", seq_len(nrow(small_mat)))
colnames(small_mat) <- paste0("column", seq_len(ncol(small_mat)))
```

# `plot_data`
`align_gg()`/`ggalign()` requires the specific data format for its operations.
If you need to transform or filter data for individual `geoms`, you can use the
`data` argument within each `geom`. However, if you have multiple `geoms` and
want a consistent transformation applied across all, you can utilize the plot
action argument `data` in the `align_gg()`/`ggalign()` function. This allows you
to transform the default data for all subsequent geoms. 

```{r}
set.seed(1234L)
ggheatmap(small_mat) +
    scale_fill_viridis_c(guide = "none") +
    anno_top() +
    align_kmeans(3L) +
    ggalign() +
    plot_data(~ subset(.x, .panel == 1L)) +
    geom_bar(aes(y = value, fill = .row_names), stat = "identity")
```

# `plot_align`
The `plot_align()` function defines the behavior of plots within a layout. It
can be used in the `action` argument of layout functions like
`quad_anno()`/`quad_active` or `stack_switch()`/`stack_active()` to set global
actions for all plots in the layout. Additionally, `plot_align()` can be applied
directly to specific plots through the `action` argument in the
`align_*()`/`ggfree()` functions, or it can be added directly to a plot.

## guides
By default, `ggheatmap()` will collect all guide legends on the side from which
they originate. 

```{r fig.dim = c(12, 12)}
heatmap_collect_all_guides <- ggheatmap(small_mat, width = 2, height = 2L) +
    scale_fill_viridis_c(name = "I'm from heatmap body") +
    theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
    anno_top() +
    align_dendro(aes(color = branch), k = 3L) +
    scale_color_brewer(
        name = "I'm from top annotation", palette = "Dark2",
        guide = guide_legend(position = "right")
    ) +
    anno_left() +
    align_dendro(aes(color = branch), k = 3L) +
    scale_color_brewer(
        name = "I'm from left annotation", palette = "Dark2",
        guide = guide_legend(position = "top", direction = "vertical")
    ) +
    anno_bottom() +
    align_dendro(aes(color = branch), k = 3L) +
    scale_color_brewer(
        name = "I'm from bottom annotation", palette = "Dark2",
        guide = guide_legend(position = "left")
    ) +
    anno_right() +
    align_dendro(aes(color = branch), k = 3L) +
    scale_color_brewer(
        name = "I'm from right annotation", palette = "Dark2",
        guide = guide_legend(position = "bottom", direction = "vertical")
    ) &
    theme(plot.margin = margin())
heatmap_collect_all_guides
```

Use the `guides` argument control which side of guide legends should be
gathered. In the following example, we'll collect the guide legends only on the
`top` and `bottom` sides: 
```{r fig.dim = c(12, 12)}
heatmap_collect_all_guides +
    # reset the active context to the heatmap layout
    quad_active() -
    # we only collect guides in the top and bottom side
    plot_align(guides = "tb")
```

You can also apply the `plot_align()` function directly to specific plots:
```{r fig.dim = c(12, 12)}
heatmap_collect_all_guides +
    # reset the active context to the heatmap layout
    quad_active() -
    # we only collect guides in the top and bottom side
    plot_align(guides = "tb") +
    # for the heatmap body, we collect guide in the right side
    plot_align(guides = "r")
```

As previously mentioned, the annotation is represented as a stack_layout()
object. The guide legends within the annotation stack are first collected by the
stack_layout() itself and then passed to the overall heatmap layout for further
integration. By default, it inherits the guides arguments from the heatmap
layout. See following example: 
```{r fig.dim = c(12, 12)}
heatmap_collect_all_guides +
    # reset the active context to the heatmap layout
    quad_active() -
    # we only collect guides in the top and bottom side
    plot_align(guides = "tb") +
    # we ensure the active context is in the bottom annotation
    # By default, it inherits "guides" argument from the heamtap layout, which
    # means it'll collect "guides" in the top and bottom side
    anno_bottom() +
    # for the dendrogram in the bottom annotation, we collect guide in the left side
    plot_align(guides = "l")
```

Here, the guide legend is collected by the bottom annotation but will not be
collected by the heatmap layout since the heatmap layout only gathers guides
from the `top` and `bottom`. In this way, the guide legends of the annotation
stack will be put around the annotation stack layout.

To override this guide collection behavior for the heatmap annotation, you can
use the `free_guides` argument of the `quad_anno()` function. This differs from the
`guides` argument in `plot_align()`, which controls the behavior for the plots
in the layout. The `free_guides` argument specifies which guide legends from the
annotation stack layout should be collected by the heatmap layout. 
```{r fig.dim = c(12, 12)}
heatmap_collect_all_guides +
    # reset the active context to the heatmap layout
    quad_active() -
    # we only collect guides in the top and bottom side
    plot_align(guides = "tb") +
    # we also collect guides in the left side for the bottom annotation stack
    # layout in the heatmap layout
    anno_bottom(free_guides = "l") +
    # for the dendrogram in the bottom annotation, we collect guide in the left side
    plot_align(guides = "l")
```

>Note: The heatmap layout will only collect guide legends from the annotation
stack if the stack layout collects its own guides first. 

## free_spaces
By default, `ggheatmap()` will align all elements of the plot, which can
sometimes lead to unwanted spacing. Consider the following example:

```{r}
ggheatmap(small_mat) +
    scale_fill_viridis_c() +
    anno_top(size = unit(30, "mm")) +
    align_dendro() +
    scale_y_continuous(
        expand = expansion(),
        labels = ~ paste("very very long labels", .x)
    ) +
    anno_left(unit(20, "mm")) +
    align_dendro()
```

In this case, the left annotation stack is positioned far from the heatmap body
due to the wide axis labels in the top annotation stack. This occurs because the
top annotation stack is aligned with the heatmap body. To fix this, you can
remove the left borders around the panel of the top annotation stack by setting
`free_spaces = "l"`.

```{r}
ggheatmap(small_mat) +
    scale_fill_viridis_c() +
    anno_top(size = unit(30, "mm")) -
    plot_align(free_spaces = "l") +
    align_dendro() +
    scale_y_continuous(
        expand = expansion(),
        labels = ~ paste("very long labels", .x)
    ) +
    anno_left(unit(20, "mm")) +
    align_dendro()
```

One useful way to utilize `free_spaces` is to position the guide legends next to
the annotations. (Note the guide legend from the bottom annotation): 
```{r fig.dim = c(12, 12)}
heatmap_collect_all_guides +
    # we only collect guides in the top and bottom side
    quad_active() -
    plot_align(guides = "tb") +
    # 1. in the bottom annotation stack layout, we collect the legends in the
    #    left side
    # 2. we remove the spaces of the left border in the annotation
    anno_bottom() -
    plot_align(guides = "l", free_spaces = "l")
```

In `ggheatmap()` and `quad_layout()`, the behavior of the `free_spaces` and
`free_labs` arguments differs from other arguments in `plot_align()` when
inheriting from the parent layout:
  - For `top` and `bottom` annotations, it inherits from the left ("l") and
  right ("r") axes.
  - For `left` and `right` annotations, it inherits from the top ("t") and
  bottom ("b") axes.

## free_labs
By default, `ggheatmap()` won't align the axis titles.
```{r}
ggheatmap(small_mat) +
    scale_fill_viridis_c() +
    ylab("Heatmap title") +
    anno_top(size = unit(30, "mm")) +
    align_dendro() +
    ylab("Annotation title")
```

To align axis titles, you can set `free_labs = NULL`. Alternatively, A single
string containing one or more of axis positions ("t", "l", "b", "r") to indicate
which axis titles should be free from alignment.

```{r}
ggheatmap(small_mat) -
    plot_align(free_labs = NULL) +
    scale_fill_viridis_c() +
    ylab("Heatmap title") +
    anno_top(size = unit(30, "mm")) +
    align_dendro() +
    ylab("Annotation title")
```

## Session information
```{r}
sessionInfo()
```
