---
title: "Plot composer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plot composer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ggalign)
```

# align_plots
Special thanks to the `patchwork` project—many core codes of the plot composer
process were adapted from `patchwork`. We have added new features to better
implement `ggalign`'s layout functions (`ggheatmap()` and `ggstack()`),
including: 

  - `free_align()`
  - `free_border()`
  - `free_guide()`
  - `free_lab()`
  - `free_space()`
  - `free_vp()`

These features have not been pushed to `patchwork` because they required
significant modification of core code. We attempted to merge them, but the
author of `patchwork` decided to implement some of these features independently.
The latest version of patchwork now includes `free_align()`, `free_lab()`, and
`free_space()` functionality under a single function: `patchwork::free()`. For
more details, see: https://www.tidyverse.org/blog/2024/09/patchwork-1-3-0/.  

The plot composer function in `ggalign` is `align_plots()`, which behaves
similarly to `cowplot::align_plots()` and `patchwork::wrap_plots()`. 

## Guide legends
By default, `align_plots()` won't collect any guide legends. You can use the
`guides` argument to control which positions of guide legends should be
collected. They will be collected to their original side. Here, we use
`patch_titles()` to indicate the guide legend position (instead of using
`ggtitle()`). `patch_titles()` can add titles on four sides, and the title will
be placed between the plot panel and the guide legend.

```{r}
p_right <- ggplot(mtcars) +
  geom_point(aes(hp, wt, colour = mpg)) +
  patch_titles("right") +
  labs(color = "right")
p_top <- p_right +
  patch_titles("top") +
  scale_color_continuous(
    name = "top",
    guide = guide_colorbar(position = "top")
  )
p_left <- p_right +
  patch_titles("left") +
  scale_color_continuous(
    name = "left",
    guide = guide_colorbar(position = "left")
  )
p_bottom <- p_right +
  patch_titles("bottom") +
  scale_color_continuous(
    name = "bottom",
    guide = guide_colorbar(position = "bottom")
  )
align_plots(p_right, p_bottom, p_top, p_left, guides = "tlbr")
```

If `align_plots()` is nested in another `align_plots()`, the nested
`align_plots()` will inherit the `guides` argument.


## free_guide
The `free_guide()` function allows you to override the `guides` argument for a
single plot.
```{r}
align_plots(
  free_guide(p_right, NULL),
  free_guide(p_bottom, NULL),
  free_guide(p_top, NULL),
  free_guide(p_left, NULL),
  guides = "tlbr"
)
```

You can also specify which guide positions to collect for individual plots.
```{r}
align_plots(
  free_guide(p_right, "r"),
  free_guide(p_bottom, "b"),
  free_guide(p_top, "t"),
  free_guide(p_left, "l")
)
```



## 
By default, `ggalign` will collect all guide legends on the side they originate
from. 

```{r setup_data}
set.seed(123)
small_mat <- matrix(rnorm(81), nrow = 9)
rownames(small_mat) <- paste0("row", seq_len(nrow(small_mat)))
colnames(small_mat) <- paste0("column", seq_len(ncol(small_mat)))
ggheatmap(small_mat) +
  scale_fill_viridis_c(name = "I'm from heatmap body") +
  theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
  hmanno("t") +
  align_dendro(aes(color = branch), k = 3L) +
  scale_color_brewer(
    name = "I'm from top tree", palette = "Dark2",
    guide = guide_legend(position = "right")
  ) +
  hmanno("l") +
  align_dendro(aes(color = branch), k = 3L) +
  scale_color_brewer(
    name = "I'm from left tree", palette = "Dark2",
    guide = guide_legend(position = "top")
  ) +
  scale_x_reverse() +
  hmanno("b") +
  align_dendro(aes(color = branch), k = 3L) +
  scale_color_brewer(
    name = "I'm from bottom tree", palette = "Dark2",
    guide = guide_legend(position = "left")
  ) +
  scale_y_reverse() +
  hmanno("r") +
  align_dendro(aes(color = branch), k = 3L) +
  scale_color_brewer(
    name = "I'm from right tree", palette = "Dark2",
    guide = guide_legend(position = "bottom")
  ) &
  theme(plot.margin = margin())
```
