---
title: "Scales and Facets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scales and Facets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

```{r setup}
library(ggalign)
```

```{r setup_data}
set.seed(123)
small_mat <- matrix(rnorm(81), nrow = 9)
rownames(small_mat) <- paste0("row", seq_len(nrow(small_mat)))
colnames(small_mat) <- paste0("column", seq_len(ncol(small_mat)))
```

## Scales
Scales are handled as the usual in ggplot2, but for position scales in the
heatmap body or in the specific axis of `align_*()` plots (referring to  the
x-axis for vertical stacks or the y-axis for horizontal stacks), we introduce
some adjustments to better align with the layout concept. 

### breaks
`breaks` and `labels` are typically handled in a similar manner of discrete
scale, thought you can provide as a continuous scale.

`breaks` should be one of:

    - `NULL` for no breaks

    - `waiver()` for the default breaks (the full data index or `NULL` if no
    data names and `labels` is `waiver()`)

    - A character vector of breaks (rownames / colunames of the matrix).

    - A numeric vector of data index (floating number were not allowed).

    - A function that takes the data limits or the data index as input and
    returns breaks as output. Also accepts rlang lambda function notation. 

```{r}
ggheatmap(small_mat) + scale_x_continuous(breaks = NULL)
```

```{r}
ggheatmap(small_mat) + scale_x_continuous()
```

```{r}
no_names <- small_mat
colnames(no_names) <- NULL
ggheatmap(no_names) + scale_x_continuous()
```

```{r}
ggheatmap(small_mat) + scale_x_continuous(breaks = c("column3", "column5"))
```

```{r}
ggheatmap(small_mat) + scale_x_continuous(breaks = c(3, 5))
```

Floating number were not allowed.
```{r error=TRUE}
ggheatmap(small_mat) + scale_x_continuous(breaks = c(3.5, 5))
```

### labels
`labels` should be one of:

    - `NULL` for no labels

    - `waiver()` for the default labels (data names)

    - A character vector giving labels (must be same length as breaks)

    - An expression vector (must be the same length as breaks). See `?plotmath`
    for details. 

    - A function that takes the data names (or breaks if data has no names) as
    input and returns labels as output. This can be also a rlang lambda
    function.

```{r}
ggheatmap(small_mat) + scale_x_continuous(labels = NULL)
```

```{r}
ggheatmap(small_mat) + scale_x_continuous()
```

```{r}
ggheatmap(small_mat) +
    scale_x_continuous(labels = letters[seq_len(ncol(small_mat))])
```

```{r}
ggheatmap(small_mat) +
    scale_x_continuous(breaks = c(3, 5), labels = c("a", "b"))
```

### data ordering
Both `breaks` and `labels` should be provided in the original scale of the raw
matrix, the internal will reorder them accordingly if you reorder the heatmap
rows/columns.

`align_reorder` will reorder the heatmap rows/columns based on the weights
returned from `order` function, or you can directly provide the ordering integer
or character index. Here we calculate the final index by the means and reorder
the heatmap columns.

We provide the labels in the original order here, the layout will reorder them.
```{r}
index <- order(colMeans(small_mat))
xlabels <- letters[seq_len(ncol(small_mat))]
print(xlabels[index])

ggheatmap(small_mat) +
    scale_x_continuous(labels = xlabels) +
    hmanno("t") +
    align_reorder(index)
```

### expand
By default, we utilize zero expand for the heatmap or for the specific axis of
`align_*()` plots (meaning the x-axis for vertical stacks or the y-axis for
horizontal stacks). This is typically the desired setting. If you wish to
introduce expansion, you must manually adjust it and apply it to each plot to
ensure proper axis alignment.
```{r}
ggheatmap(small_mat) +
    scale_x_continuous(expand = expansion(mult = 0.1)) +
    hmanno("t") +
    align_dendro(aes(color = branch), k = 3L) +
    scale_x_continuous(expand = expansion(mult = 0.1))
```

## Facets
When working with facets, manual configuration of panel using the `facet_*()`
function is not possible since the internal will use `facet_grid` to set the
row/column groups defined by `align_*()` functions. However, you can provide
`facet_grid()` or `facet_null()` (if no panel) to control other arguments except
`rows` or `cols`. A common usage case is to change the panel strip text.

The default theme in this package will always remove the panel strip text, you
can override this behaviour with `theme(strip.text = element_text())`. This
allows us to add the panel title in the plot area.
```{r}
ggheatmap(small_mat) +
    facet_grid(labeller = labeller(.column_panel = function(x) letters[as.integer(x)])) +
    theme(strip.text = element_text()) +
    hmanno("top") +
    align_kmeans(centers = 3L)
```

## Coords
Currently, only cartesian coordinate can be used to align axis well.
Internally, the limits will always be set to the number of observations, with an
additional range expansion of 0.5 added on both ends. 

## Session information
```{r}
sessionInfo()
```
