---
title: "Scales and Facets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scales and Facets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ggalign)
```

```{r setup_data}
set.seed(123)
small_mat <- matrix(rnorm(81), nrow = 9)
rownames(small_mat) <- paste0("row", seq_len(nrow(small_mat)))
colnames(small_mat) <- paste0("column", seq_len(ncol(small_mat)))
```

## Scales
Scales are handled as the usual in ggplot2, but for position scales, `limits`
cannot be setted, and the internal will always reset it as the default. This is
required to align the layout and annotation.
```{r}
ggheatmap(small_mat) + scale_x_continuous(limits = c(0, 0))
```

`breaks` and `labels` in the position scales are handled in a similar manner of
discrete scale, thought you can provide as a continuous scale.

### breaks
`breaks` should be one of:

    - `NULL` for no breaks

    - `waiver()` for the default breaks (the full data index or `NULL` if no
    data names and `labels` is `waiver()`)

    - A character vector of breaks (rownames / colunames of the matrix).

    - A numeric vector of data index.

    - A function that takes the data limits or the data index as input and
    returns breaks as output. Also accepts rlang lambda function notation. 

```{r}
ggheatmap(small_mat) + scale_x_continuous(breaks = NULL)
```

```{r}
ggheatmap(small_mat) + scale_x_continuous()
```

```{r}
no_names <- small_mat
colnames(no_names) <- NULL
ggheatmap(no_names) + scale_x_continuous()
```

```{r}
ggheatmap(small_mat) + scale_x_continuous(breaks = c("column3", "column5"))
```

```{r}
ggheatmap(small_mat) + scale_x_continuous(breaks = 5:6)
```

### labels
`labels` should be one of:

    - `NULL` for no labels

    - `waiver()` for the default labels (data names)

    - A character vector giving labels (must be same length as breaks)

    - An expression vector (must be the same length as breaks). See `?plotmath`
    for details. 

    - A function that takes the data names (or breaks if data has no names) as
    input and returns labels as output. This can be also a rlang lambda
    function.

```{r}
ggheatmap(small_mat) + scale_x_continuous(labels = NULL)
```

```{r}
ggheatmap(small_mat) + scale_x_continuous()
```

```{r}
ggheatmap(small_mat) +
  scale_x_continuous(labels = letters[seq_len(ncol(small_mat))])
```

```{r}
ggheatmap(small_mat) +
  scale_x_continuous(breaks = c(3, 5), labels = c("a", "b"))
```

### discrete or continuous
`breaks` and `labels` argument will work the same in both discrete scale and
continuous scale. But note it's hard to match all discrete scale limits, for
sometimes, you should manually specify the `expand` argument.

I will provide an example of using a heatmap annotation. If you are unfamiliar
with adding heatmap annotations, try to see `vignette("layout-heatmap")`. If you
want to learn more about the `align_*` functions, try to see
`vignette("align-layout")` and `vignette("align-plot")`. 

`hmanno()` indicates where the annotation was added. `align_dendro()` will
reorder the layout based on the hierarchical clustering tree, and it'll also
draw the dendrogram tree.

In following example, we must add `expand` argument to match the heatmap and
the tree.
```{r}
ggheatmap(small_mat) +
  scale_x_discrete(breaks = c(3, 5), labels = c("a", "b")) +
  hmanno("t") +
  align_dendro() +
  scale_x_discrete()
```

```{r}
ggheatmap(small_mat) +
  scale_x_discrete(breaks = c(3, 5), labels = c("a", "b")) +
  hmanno("t") +
  align_dendro() +
  scale_x_discrete(expand = expansion(add = 0.5))
```

Indeed, you can use the default scale for the dendrogram directly. The layout
always sets the appropriate limits for the continuous scale.

```{r}
ggheatmap(small_mat) +
  scale_x_discrete(breaks = c(3, 5), labels = c("a", "b")) +
  hmanno("t") +
  align_dendro()
```

One useful function when you should use the discrete scale is the ambiguous
`orientation` in geoms. Some geoms use the discrete axis as the main orientation
like `ggplot2::geom_bar()` and `ggplot2::geom_box()`, but by default we always
use the continuous scale. So you must manually provide an `orientation` argument
for these geoms or manually set the discrete scale in the horizontal layout
(including left and right heatmap annotation).

`ggalign()` is similar to `ggplot` in that it initializes a plot data and
`mapping`. The data input can be a matrix, a data frame, or a simple vector that
will be converted into a one-column matrix, and can inherit from the layout
data. When a plot is initialized, we can add other ggplot2 elements for it. You
can see `ggplot2 specification` in `?align_gg` for details.

In the following example, `geom_bar` won't work since the ambiguous orientation.
```{r}
ggheatmap(small_mat) +
  hmanno("l") +
  ggalign() +
  geom_bar(aes(x = value), stat = "identity") +
  theme(axis.text.x = element_text(angle = -60, hjust = 0))
```

Try to provide the `orientation` argument or set the scale into discrete.
```{r}
ggheatmap(small_mat) +
  hmanno("l") +
  ggalign() +
  geom_bar(aes(x = value), stat = "identity", orientation = "y") +
  theme(axis.text.x = element_text(angle = -60, hjust = 0))
```

```{r}
ggheatmap(small_mat) +
  hmanno("l") +
  ggalign() +
  geom_bar(aes(x = value), stat = "identity") +
  scale_y_discrete() +
  theme(axis.text.x = element_text(angle = -60, hjust = 0))
```

### argument ordering
Both `breaks` and `labels` should be provided in the original scale of the raw
matrix, the internal will reorder them accordingly if you reorder the heatmap
rows/columns.

`align_reorder` will reorder the heatmap rows/columns based on the weights
returned from `fun` argument. Here we reorder the heatmap columns by the means.

We provide the labels in the original order here, the layout will reorder them.
```{r}
index <- order(colMeans(small_mat))
xlabels <- letters[seq_len(ncol(small_mat))]
print(xlabels[index])

ggheatmap(small_mat) +
  scale_x_continuous(labels = xlabels) +
  hmanno("t") +
  align_reorder()
```

## Facets
When working with facets, manual configuration of panel using the `facet_*`
function is not possible since the internal will use `facet_grid` to set the
row/column groups specified by the heatmap annotation. However, you can provide
`facet_grid()` or `facet_null()` (if no panel) to control other arguments except
`rows` and `cols`. A common usage case is to change the panel strip text.

The default theme in this package will always remove the panel strip text, you
can override this behaviour with `theme(strip.text = element_text())`. This
allows us to add the panel title in heatmaps or any heatmap annotation with
plot.
```{r}
ggheatmap(small_mat) +
  facet_grid(labeller = labeller(.column_panel = function(x) letters[as.integer(x)])) +
  theme(strip.text = element_text()) +
  hmanno("top") +
  align_kmeans(centers = 3L)
```

## Session information
```{r}
sessionInfo()
```
