---
title: "Stack Layout"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Stack Layout}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

`stack_layout()` arranges plots either horizontally or vertically, and can be
referred to using the alias `ggstack()`. There are two types of stack layouts:

  - `stack_align()`: align the observations along the stack.
  - `stack_free()`: don't align the observations.

Several aliases are available for convenience:

  - `stack_alignv`: Aligns the stack vertically (special case of `stack_align()`).
  - `stack_alignh`: Aligns the stack horizontally (special case of `stack_align()`).
  - `stack_freev`: A vertical version of `stack_free()`.
  - `stack_freeh`: A horizontal version of `stack_free()`.
  `direction = "horizontal"`.

```{r setup}
library(ggalign)
```

# Input data 

- For `stack_free()`, a data frame is required, and the input will be automatically
  converted using `fortify_data_frame()` if needed. 
- For `stack_align()`, a matrix is required, and the input will be automatically
  converted using `fortify_matrix()` if needed. 

```{r setup_data}
set.seed(123)
small_mat <- matrix(rnorm(56), nrow = 7)
rownames(small_mat) <- paste0("row", seq_len(nrow(small_mat)))
colnames(small_mat) <- paste0("column", seq_len(ncol(small_mat)))
```

By default, `stack_align()`/`stack_free()` will set up the layout, but no plot
will be drawn until you add a plot element:
```{r}
stack_alignh(small_mat)
# stack_freev(small_mat) # the same for `stack_free()`
```

# plot element
For `stack_align()`, plots can be added regardless of whether they need to
align observations.
```{r}
stack_alignh(small_mat) + align_dendro()
```

```{r}
stack_alignh(small_mat) +
    align_kmeans(centers = 3L) +
    ggalign(data = rowSums) +
    geom_bar(aes(value, fill = .panel), orientation = "y", stat = "identity") +
    facet_grid(switch = "y") +
    theme(strip.text = element_text()) +
    align_dendro(aes(color = branch))
```

We can stack the plots vertically with `stack_alignv()`:
```{r}
stack_alignv(small_mat) + align_dendro()
```

>Note that vertical stack take the `x-axis` as the observations, but horizontal
stack take the `y-axis` as the observations.

`stack_align()` can add plot without alignment of observations.
```{r}
stack_alignv() +
    ggfree(aes(displ, hwy, colour = class), data = mpg) +
    geom_point(size = 2) +
    ggfree(aes(displ, hwy, colour = class), data = mpg) +
    geom_point(size = 2) &
    scale_color_brewer(palette = "Dark2") &
    theme_bw()
```

For `stack_free()`, only plots that are free from alignment (`ggfree()`) can be
added along the axis. `stack_free()` alone is not very useful, since we can
combine multiple plots vertically or horizontally with `align_plots()` or
[patchwork](https://cran.r-project.org/package=patchwork). The only benefits
would be the data can be inherited by multiple plots.
```{r}
stack_freev(mpg) +
    ggfree(aes(displ, hwy, colour = class)) +
    geom_point(size = 2) +
    ggfree(aes(displ, hwy, colour = class)) +
    geom_point(size = 2) &
    scale_color_brewer(palette = "Dark2") &
    theme_bw()
```

# `ggheatmap()`/`quad_layout()`
`ggheatmap()`/`quad_layout()` can be added to a stack layout. 
```{r}
stack_alignh(small_mat) +
    ggheatmap()
```

However, as noted earlier, `stack_free()` can only accept plots that are free
from alignment along the axis. Here's a table summarizing which layouts are
compatible with `stack_free()`:

|                               | `stack_freeh()` | `stack_freev()` |
| ----------------------------- | --------------- | --------------- |
| `quad_free()`/`ggside()`      | ✅              | ✅              |
| `quad_alignh()`               | ❌              | ✅              |
| `quad_alignv()`               | ✅              | ❌              |
| `quad_alignb()`/`ggheatmap()` | ❌              | ❌              |

```{r}
alignh_quad_plot <- quad_alignh(small_mat) +
    geom_boxplot(aes(value, .row_names, fill = .row_names)) +
    scale_fill_brewer(palette = "Dark2")
stack_freev() +
    alignh_quad_plot +
    alignh_quad_plot
```

The data of `ggheatmap()`/`quad_layout()` can inherit from the parent
stack_layout(), but the data format must match. Otherwise, you will need to
manually provide the data. When `ggheatmap()`/`quad_layout()` is added to a
vertical stack_align(), the inherited matrix is transposed before use. 
```{r}
stack_alignv(small_mat) +
    ggheatmap()
```

Once a `ggheatmap()`/`quad_layout()` is added, any further elements you add will
be applied to this `ggheatmap()`/`quad_layout()`. You can include `align_*`
elements or any ggplot2 components for the `quad_layout()`.
```{r}
stack_alignh(small_mat) +
    ggheatmap() +
    scale_fill_viridis_c()
```

If you'd like to add elements to the stack layout rather than the
`ggheatmap()`/`quad_layout()` layout, you can easily switch from the
`ggheatmap()`/`quad_layout()` to the stack layout using `stack_active()`. 
```{r}
stack_alignh(small_mat) +
    ggheatmap() +
    scale_fill_viridis_c() +
    theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
    stack_active() +
    ggalign(data = rowSums) +
    geom_bar(aes(value), fill = "red", orientation = "y", stat = "identity")
```

One exception is the `ggheatmap()`/`quad_layout()` itself, which cannot be added
to another heatmap layout. Therefore, you can directly add a
`ggheatmap()`/`quad_layout()` to the stack layout without using
`stack_active()`.
```{r}
stack_alignv(small_mat) +
    ggheatmap() +
    ggheatmap() &
    scale_fill_viridis_c()
```

# Control sizes
A length of 3 sizes should be provided in `stack_align()`/`stack_free()` when
putting a heatmap with flank annotation into the stack layout. For vertical
stacks, this means heatmaps with left or right annotations; for horizontal
stacks, this means heatmaps with top or bottom annotations. The first size
controls the relative `width`/`height` of the left or top annotation, the second
controls the relative size of the heatmap body `width`/`height`, and the third
controls the relative `width`/`height` of the right or bottom annotation. 

By default the three rows/columns will have equal sizes.
```{r}
stack_alignv(small_mat) +
    ggheatmap() +
    scale_fill_viridis_c() +
    theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
    anno_left() +
    align_dendro(aes(color = .panel), k = 3L) +
    anno_right() +
    ggalign(data = rowSums) +
    geom_bar(aes(value, fill = .panel), orientation = "y", stat = "identity") +
    ggheatmap() +
    theme(axis.text.x = element_text(angle = -60, hjust = 0))
```

```{r}
# the left annotation, will have a relative size 1
# the heatmap body will have a relative width 2
# the right annotation will have a relative size 1
stack_alignv(small_mat, sizes = c(1, 2, 1)) +
    ggheatmap() +
    scale_fill_viridis_c() +
    theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
    anno_left() +
    align_dendro(aes(color = .panel), k = 3L) +
    anno_right() +
    ggalign(data = rowSums) +
    geom_bar(aes(value, fill = .panel), orientation = "y", stat = "identity") +
    ggheatmap() +
    theme(axis.text.x = element_text(angle = -60, hjust = 0))
```

In this way, heatmap body width/height specified in `quad_active()` or
`ggheatmap()` won't work.
```{r}
stack_alignv(small_mat) +
    ggheatmap(width = unit(2, "null")) + # not work
    scale_fill_viridis_c() +
    theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
    anno_left() +
    align_dendro(aes(color = .panel), k = 3L) +
    anno_right() +
    ggalign(data = rowSums) +
    geom_bar(aes(value, fill = .panel), orientation = "y", stat = "identity") +
    ggheatmap(width = unit(2, "null")) + # not work
    theme(axis.text.x = element_text(angle = -60, hjust = 0))
```

By default the flank annotation will fill the whole stack flank, but we can
still control the size of heatmap annotation in `quad_anno()`. 
```{r}
stack_alignv(small_mat) +
    ggheatmap() +
    scale_fill_viridis_c() +
    theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
    # the left annotation will have a total size 2cm
    anno_left(size = unit(2, "cm")) +
    align_dendro(aes(color = .panel), k = 3L) +
    anno_right() +
    ggalign(data = rowSums) +
    geom_bar(aes(value, fill = .panel), orientation = "y", stat = "identity") +
    ggheatmap() +
    theme(axis.text.x = element_text(angle = -60, hjust = 0))
```

You can also use `npc` unit.
```{r}
stack_alignv(small_mat) +
    ggheatmap() +
    scale_fill_viridis_c() +
    theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
    # the left annotation will have a total width 1/2 npc
    anno_left(size = unit(0.5, "npc")) +
    align_dendro(aes(color = .panel), k = 3L) +
    anno_right() +
    ggalign(data = rowSums) +
    geom_bar(aes(value, fill = .panel), orientation = "y", stat = "identity") +
    ggheatmap() +
    theme(axis.text.x = element_text(angle = -60, hjust = 0))
```

## Session information
```{r}
sessionInfo()
```
