---
title: "FAQ"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{FAQ}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

```{r setup}
library(ggalign)
```

```{r setup_data}
set.seed(123)
small_mat <- matrix(rnorm(81), nrow = 9)
rownames(small_mat) <- paste0("row", seq_len(nrow(small_mat)))
colnames(small_mat) <- paste0("column", seq_len(ncol(small_mat)))
```

## How to customize the gene ordering in ggoncoplot function?
To manually define the row (gene) and column (sample) order in `ggoncoplot()`,
you first need to disable the automatic ordering applied by `ggoncoplot()`.
Then, you can use the `align_order()` function to specify the desired
arrangement.

```{r}
# Sample data
mut_mat <- read.table(
    textConnection(
        "s1,s2,s3
         snv;indel,snv,indel
         ,snv;indel,snv
         snv,,indel;snv"
    ),
    row.names = NULL,
    header = TRUE,
    sep = ",",
    stringsAsFactors = FALSE
)
rownames(mut_mat) <- c("g1", "g2", "g3")
mut_mat <- as.matrix(mut_mat)

# Example 1: Manually order rows using numeric indices
ggoncoplot(
    mut_mat,
    map_width = c(snv = 0.5),
    map_height = c(indel = 0.9),
    reorder_row = FALSE
) +
    scale_fill_brewer(palette = "Dark2", na.translate = FALSE) +
    # Use anno_left() or anno_right() for row ordering; for columns, use anno_top() or anno_bottom()
    anno_left() +
    align_order(c(2, 3, 1)) # Orders rows as g2, g3, g1
```

```{r}
# Example 2: Manually order rows using row names
ggoncoplot(
    mut_mat,
    map_width = c(snv = 0.5),
    map_height = c(indel = 0.9),
    reorder_row = FALSE
) +
    scale_fill_brewer(palette = "Dark2", na.translate = FALSE) +
    anno_left() +
    align_order(c("g2", "g1", "g3")) # Orders rows as g2, g1, g3
```

For row ordering, you can use either numeric indices (e.g., c(2, 3, 1)) or
character names (e.g., c("g2", "g1", "g3")) if your matrix has row names. For
column ordering, simply replace `anno_left()` with `anno_top()` or
`anno_bottom()` and specify the sample order in `align_order()`. Note the
ordering specified is from bottom to the top or from left to the right which
following the ggplot2 coordinates.

For certain objects related to `ggoncoplot()`, the `fortify_matrix()` method
includes a `genes` argument that allows manual specification of gene order
(character only). You can define this argument directly within `ggoncoplot()`,
Currently, the supported objects for oncoplot-specific usage include:

 - `maftools`
 - a matrix wrapped with `tune(matrix, "oncoplot")`

The primary reason we must wrap a standard matrix with `tune()` is that
`fortify_matrix()` cannot distinguish between the default matrix method and the
one specifically intended for `ggoncoplot()`.

```{r}
ggoncoplot(
    tune(mut_mat, "oncoplot"),
    genes = c("g2", "g1", "g3"), # Orders rows as g2, g1, g3
    map_width = c(snv = 0.5),
    map_height = c(indel = 0.9),
    reorder_row = FALSE
) +
    scale_fill_brewer(palette = "Dark2", na.translate = FALSE) 
```

## Can I change the default data for all geoms?
All function in `ggalign` requires specific data formats for its operations. If
you need to transform or filter data for individual `geoms`, you can use the
`data` argument within each `geom`. However, if you have multiple `geoms` and
want a consistent transformation applied across all, you can utilize the
`scheme_data()` function which allows you to transform the default data for all
additive geoms.

## Why are there spaces after setting all `plot.margin` to zero?
By default, ggalign doesn't add spaces between plots. If you notice spaces, they
are likely due to:

- Plot margins within individual plots.
- Scale expansion automatically added by ggplot2.

You can resolve this by removing the scale expansion and adjusting the margins
in the theme: 
```{r}
ggheatmap(small_mat) +
    anno_top() +
    align_dendro(aes(color = branch), k = 3L) +
    scale_y_continuous(expand = expansion()) &
    theme(plot.margin = margin())
```

## Why can't I add two dendrograms with different `reorder` settings?

> Note: We always prevent users from reordering layout direction twice.

When `reorder_group = FALSE` is used, reordering of the heatmap occurs within
each group. As long as the ordering within each group remains consistent, these
two dendrograms can be placed on the same axis of the heatmap. 

Let's assume that panels represents the group for each observation and index
represents the ordered index of the observations. 

In the code provided, panels is initialized as ungrouped and index as unordered: 
```{r}
panels <- NULL
index <- NULL
```

```{r}
set.seed(2L)
group <- sample(letters[1:3], ncol(small_mat), replace = TRUE)
h <- ggheatmap(small_mat) +
    anno_top() +
    align_group(group)
```

When we add a dendrogram with `reorder_group = FALSE`; this'll do something
like this (Since `reorder_group = FALSE`, the `panels` won't be changed): 
```{r}
h1 <- h + align_dendro(reorder_group = FALSE, merge_dendrogram = TRUE)
# always remember dendrogram will initialize the index, here, we extract the
# index from the underlying dendrogram
index <- stats::order.dendrogram(ggalign_stat(h1, "top", 2L))
```

However, if a dendrogram with `reorder_group = TRUE` is then added, it will
reorder the panels, so the underlying index will be changed: 
```{r}
h2 <- h + align_dendro(reorder_group = TRUE, merge_dendrogram = TRUE)
new_index <- stats::order.dendrogram(ggalign_stat(h2, "top", 2L))
all(index == new_index)
```

To prevent reordering of the axis twice, we ensure that the `new_index` matches
the `index` (if `index` is not NULL). This is done to maintain consistency and
prevent conflicts in the ordering of the heatmap.

## Session information
```{r}
sessionInfo()
```
