---
title: "Quad-Layout"
output: rmarkdown::html_vignette
params:
  mode: !r if (endsWith(as.character(desc::desc_get_version()), '9000')) 'devel' else 'release'
  release: "https://yunuuuu.github.io/ggalign"
  devel: "https://yunuuuu.github.io/ggalign/dev"
vignette: >
  %\VignetteIndexEntry{Quad-Layout}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

```{r echo=FALSE}
mode <- params$mode
url <- params[[mode]]
```

As discussed in 
[heatmap-layout](`r sprintf("%s/articles/heatmap-layout.html", url)`), the
`ggheatmap()` function is a specialized form of `quad_layout()`. Specifically,
it is a variant of `quad_alignb()`, which aligns observations both horizontally
and vertically.

`quad_layout()` arranges plots in the Quad-Side of a main plot. Depending on
whether you want to align observations in the `horizontal` or `vertical`
direction, there are four main types of `quad_layout()`: 
 - `quad_free`/`ggside`: Never align observations.
 - `quad_alignh`: Align observations in the horizontal direction.
 - `quad_alignv`: Align observations in the vertical direction.
 - `quad_alignb`: Align observations in both horizontal and vertical directions. 

```{r setup}
library(ggalign)
```

# `quad_free()`
`quad_free()` won't align observations. `ggside()` is an alias of `quad_free()`.
It do the same function of [ggside](https://cran.r-project.org/package=ggside)
packge. This is particularly useful for metadata for summary graphics on a
continuous axis 

## input data
Like `ggplot`, you'll need to provide input that can be converted into a data
frame. Internally, the `fortify_data_frame()` function is used to handle this
conversion. 

```{r}
ggside(mpg) +
    geom_point(aes(displ, hwy, colour = class))
```

## annotation
Like `ggheatmap()`, annotations in `quad_free()` can be placed at the top, left,
bottom, and right sides, but all annotations here use `stack_free()`, meaning
they don't align observations. The data will be directly inherited which is
different from the `quad_alignb()`/`quad_alignh()`/`quad_alignv()` where the
column annotation will transpose the layout matrix 

Additionally, annotations can be manually initialized with user-specified data
using `quad_init()`, I'll demonstrate it in `quad_alignh()`/`quad_alignv()`
section where manual initialization is required.

```{r}
ggside(mpg, aes(displ, hwy, colour = class)) -
    # set default theme for all plots in the layout
    plot_theme(theme_bw()) +
    geom_point(size = 2) +
    # add top annotation
    anno_top(size = 0.3) -
    # set default theme for the top annotation
    plot_theme(theme_no_axes("tb")) +
    # add a plot in the top annotation
    ggfree() +
    geom_density(aes(displ, y = after_stat(density), colour = class), position = "stack") +
    # add right annotation
    anno_right(size = 0.3) -
    # set default theme for the right annotation
    plot_theme(theme_no_axes("lr")) +
    # add a plot in the right annotation
    ggfree() +
    geom_density(aes(x = after_stat(density), hwy, colour = class),
        position = "stack"
    ) +
    theme(axis.text.x = element_text(angle = 90, vjust = .5))
```

Unlike `ggheatmap()`/`heatmap_layout()`, `ggside()`/`quad_free()` allows
facetting for the main plot, which should also be applied to the annotations for
proper alignment.
```{r}
i2 <- iris
i2$Species2 <- rep(c("A", "B"), 75)
ggside(i2, aes(Sepal.Width, Sepal.Length, color = Species)) -
    # set default theme for all plots in the layout
    plot_theme(theme_bw()) +
    geom_point(size = 2) +
    facet_grid(Species ~ Species2) +
    # add top annotation
    anno_top(size = 0.3) -
    # set default theme for the top annotation
    plot_theme(theme_no_axes("tb")) +
    # add a plot in the top annotation
    ggfree() +
    geom_density(aes(Sepal.Width, y = after_stat(density), colour = Species),
        position = "stack"
    ) -
    # we can utilize the `-` to set facet for all plots in the annotation
    facet_grid(cols = vars(Species2)) +
    # add right annotation
    anno_right(size = 0.3) -
    # set default theme for the right annotation
    plot_theme(theme_no_axes("lr")) +
    # add a plot in the right annotation
    ggfree() +
    geom_density(aes(x = after_stat(density), Sepal.Length, colour = Species),
        position = "stack"
    ) +
    theme(axis.text.x = element_text(angle = 90, vjust = .5)) -
    # we can utilize the `-` to set facet for all plots in the annotation
    facet_grid(rows = vars(Species))
```

# `quad_alignh()`/`quad_alignv()`
`quad_alignh()` aligns observations horizontally, while `quad_alignv()` aligns
observations vertically.

## input data
To align observations in the horizontal/vertical direction, you'll need to
provide input that can be converted into a matrix. Internally, the
`fortify_matrix()` function is used to handle this conversion, ensuring the
input is in the correct format for alignment.

The underlying ggplot data is the same with `ggheatmap()`/`heatmap_layout()`, it
is recommended to use `.y`, or `.row_names` as the `y` mapping in the main plot
for `quad_alignh()` but use `.x` or `.column_names` as the `x` mapping in the
main plot for `quad_alignv()`.

```{r setup_data}
set.seed(123)
small_mat <- matrix(rnorm(56), nrow = 8)
rownames(small_mat) <- paste0("row", seq_len(nrow(small_mat)))
colnames(small_mat) <- paste0("column", seq_len(ncol(small_mat)))
```

```{r}
# we'll align the horizontal direction
quad_alignh(small_mat) +
    geom_boxplot(aes(value, .row_names, fill = .row_names)) +
    scale_fill_brewer(palette = "Dark2")
```

```{r}
# we'll align the vertical direction
quad_alignv(small_mat) +
    geom_boxplot(aes(.column_names, value, fill = .column_names)) +
    scale_fill_brewer(palette = "Dark2")
```

## annotation
the same with `ggheatmap()`/`heatmap_layout()`, we can add annotation in top,
left, bottom, and right side. but the top and bottom annotation in
`quad_alignh()` is a `stack_free()` which won't align observations and left and
right annotation in `quad_alignv()` is a `stack_free()` which won't align
observations.

For left and right annotation in `quad_alignh()`, they are a `stack_align()`
layout which will align observations, and require a matrix in the layout data.
So it can inherit from the parent `quad_layout()` and we can add both
`ggalign()` and `ggfree()` and other `align_*()` functions to the annotation.
```{r}
quad_alignh(small_mat) +
    geom_boxplot(aes(value, .row_names, fill = .row_names)) +
    scale_fill_brewer(palette = "Dark2") +
    anno_left() +
    align_dendro()
```

For top and bottom annotation in `quad_alignh()`, they are a `stack_free()`
which won't align observations, and require a data frame in the layout data. In
this way, we won't transform the matrix to a data frame and always prevent user
from inheriting from the parent layout matrix.

```{r error=TRUE}
quad_alignh(small_mat) +
    geom_boxplot(aes(value, .row_names, fill = .row_names)) +
    scale_fill_brewer(palette = "Dark2") +
    anno_top()
```

Manual initialization of annotations with `quad_init()` is required in such
cases: 
```{r}
quad_alignh(small_mat) +
    geom_boxplot(aes(value, .row_names, fill = .row_names)) +
    scale_fill_brewer(palette = "Dark2") +
    # initialize the top annotation with a self-defined data frame
    quad_init("top", data = mpg) +
    # add a plot in the top annotation
    ggfree(aes(displ, hwy, colour = class)) +
    geom_point(aes(displ, hwy, colour = class))
```

The same with `quad_alignh()`, we need to manually initialize the left and right
annotations in `quad_alignv()`.
```{r}
quad_alignv(small_mat) +
    geom_boxplot(aes(.column_names, value, fill = .column_names)) +
    scale_fill_brewer(palette = "Dark2") +
    # initialize the left annotation with a self-defined data frame
    quad_init("left", data = mpg) +
    # add a plot in the top annotation
    ggfree(aes(displ, hwy, colour = class)) +
    geom_point(aes(displ, hwy, colour = class))
```

# `quad_alignb()`
`quad_alignb` aligns observations in both horizontal and vertical directions. It
is a base version of `ggheatmap()`/`heatmap_layout()`, it won't add the
main plot geoms and default mapping automatically.

# Session information
```{r}
sessionInfo()
```
