---
title: "ComplexHeatmap: single-heatmap"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ComplexHeatmap: single-heatmap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

In this thread, we'll use `ggalign` to draw all the heatmap in
<https://jokergoo.github.io/ComplexHeatmap-reference/book/a-single-heatmap.html>

```{r setup}
library(ggalign)
```

```{r}
set.seed(123)
nr1 <- 4
nr2 <- 8
nr3 <- 6
nr <- nr1 + nr2 + nr3
nc1 <- 6
nc2 <- 8
nc3 <- 10
nc <- nc1 + nc2 + nc3
mat <- cbind(
    rbind(
        matrix(rnorm(nr1 * nc1, mean = 1, sd = 0.5), nrow = nr1),
        matrix(rnorm(nr2 * nc1, mean = 0, sd = 0.5), nrow = nr2),
        matrix(rnorm(nr3 * nc1, mean = 0, sd = 0.5), nrow = nr3)
    ),
    rbind(
        matrix(rnorm(nr1 * nc2, mean = 0, sd = 0.5), nrow = nr1),
        matrix(rnorm(nr2 * nc2, mean = 1, sd = 0.5), nrow = nr2),
        matrix(rnorm(nr3 * nc2, mean = 0, sd = 0.5), nrow = nr3)
    ),
    rbind(
        matrix(rnorm(nr1 * nc3, mean = 0.5, sd = 0.5), nrow = nr1),
        matrix(rnorm(nr2 * nc3, mean = 0.5, sd = 0.5), nrow = nr2),
        matrix(rnorm(nr3 * nc3, mean = 1, sd = 0.5), nrow = nr3)
    )
)
mat <- mat[sample(nr, nr), sample(nc, nc)]
rownames(mat) <- paste0("row", seq_len(nr))
colnames(mat) <- paste0("column", seq_len(nc))
```

# Colors
It is important to note that the `ComplexHeatmap` package reorders the
dendrogram by default, while `align_dendro()` in `ggalign` does not modify the
tree layout. 

Another key difference is in how the two packages treat the starting point.
`ggalign` considers the left-bottom as the starting point, whereas
`ComplexHeatmap` starts from the left-top. When reordering the dendrogram,
`ComplexHeatmap` does so in decreasing order, while `ggalign` uses an ascending
order. 

To modify colors in the heatmap, you can use the `scale_fill_*()` function from
`ggplot2`, which provides a flexible way and enriched pallete to adjust color
schemes.

```{r}
# ComplexHeatmap::Heatmap(mat)
dim(mat)
ggheatmap(mat) +
    theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
    anno_right(size = unit(15, "mm")) +
    align_dendro(reorder_dendrogram = TRUE) +
    anno_top(size = unit(15, "mm")) +
    align_dendro(reorder_dendrogram = TRUE) +
    quad_active() &
    theme(plot.margin = margin())
```

```{r}
ggheatmap(mat) +
    scale_fill_gradient2(low = "green", high = "red") +
    theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
    anno_right(size = unit(15, "mm")) +
    align_dendro(reorder_dendrogram = TRUE) +
    anno_top(size = unit(15, "mm")) +
    align_dendro() +
    quad_active() &
    theme(plot.margin = margin())
```

`oob` argument in the `scale_fill_*` function can be used to deal with the
outliers.

```{r}
mat2 <- mat
mat2[1, 1] <- 100000
ggheatmap(mat2) +
    scale_fill_gradient2(
        low = "green", high = "red",
        limits = c(-2, 2),
        oob = scales::squish
    ) +
    theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
    anno_right(size = unit(15, "mm")) +
    align_dendro(reorder_dendrogram = TRUE) +
    theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
    anno_top(size = unit(15, "mm")) +
    align_dendro(reorder_dendrogram = TRUE) +
    quad_active() &
    theme(plot.margin = margin())
```

We can use `align_plots()` to arrange them.
```{r fig.dim = c(10, 10)}
h1 <- ggheatmap(mat) +
    scale_fill_gradient2(name = "mat", low = "green", high = "red") +
    theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
    anno_right(size = unit(15, "mm")) +
    align_dendro(reorder_dendrogram = TRUE) +
    theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
    anno_top(size = unit(15, "mm")) +
    align_dendro(reorder_dendrogram = TRUE) +
    quad_active() &
    theme(plot.margin = margin())

h2 <- ggheatmap(mat / 4) +
    scale_fill_gradient2(
        name = "mat/4", limits = c(-2, 2L),
        oob = scales::squish,
        low = "green", high = "red"
    ) +
    theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
    anno_right(size = unit(15, "mm")) +
    align_dendro(reorder_dendrogram = TRUE) +
    theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
    anno_top(size = unit(15, "mm")) +
    align_dendro(reorder_dendrogram = TRUE) +
    quad_active() &
    theme(plot.margin = margin())

h3 <- ggheatmap(abs(mat)) +
    scale_fill_gradient2(name = "abs(mat)", low = "green", high = "red") +
    theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
    anno_right(size = unit(15, "mm")) +
    align_dendro(reorder_dendrogram = TRUE) +
    theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
    anno_top(size = unit(15, "mm")) +
    align_dendro(reorder_dendrogram = TRUE) +
    quad_active() &
    theme(plot.margin = margin())
align_plots(h1, h2, h3, ncol = 2L)
```

```{r}
ggheatmap(mat) +
    scale_fill_gradientn(colors = rev(rainbow(10))) +
    theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
    anno_right(size = unit(15, "mm")) +
    align_dendro(reorder_dendrogram = TRUE) +
    anno_top(size = unit(15, "mm")) +
    align_dendro(reorder_dendrogram = TRUE) +
    quad_active() &
    theme(plot.margin = margin())
```

For character matrix, you can use ggplot2 discrete filling scales.
```{r}
discrete_mat <- matrix(sample(1:4, 100, replace = TRUE), 10, 10)
colors <- structure(1:4, names = c("1", "2", "3", "4")) # black, red, green, blue
ggheatmap(discrete_mat, aes(fill = factor(value))) +
    scale_fill_manual(values = colors) +
    anno_right(size = unit(15, "mm")) +
    align_dendro(reorder_dendrogram = TRUE) +
    anno_top(size = unit(15, "mm")) +
    align_dendro(reorder_dendrogram = TRUE) +
    quad_active() &
    theme(plot.margin = margin())
```

```{r}
discrete_mat <- matrix(sample(letters[1:4], 100, replace = TRUE), 10, 10)
colors <- structure(1:4, names = letters[1:4])
ggheatmap(discrete_mat) +
    scale_fill_manual(values = colors)
```

```{r}
mat_with_na <- mat
na_index <- sample(c(TRUE, FALSE),
    nrow(mat) * ncol(mat),
    replace = TRUE, prob = c(1, 9)
)
mat_with_na[na_index] <- NA
ggheatmap(mat_with_na) +
    scale_fill_gradient2(low = "blue", high = "red", na.value = "black") +
    theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
    anno_right(size = unit(15, "mm")) +
    align_dendro(reorder_dendrogram = TRUE) +
    anno_top(size = unit(15, "mm")) +
    align_dendro(reorder_dendrogram = TRUE) +
    quad_active() &
    theme(plot.margin = margin())
```

We won't compare the LAB and RGB space. If you want to convert color between
different color space, try to use [farver](https://farver.data-imaginist.com/)
pacakge.

In ggplot2, you can use `panel.border` argument in `theme()` function to control
the Heatmap body border.
```{r}
ggheatmap(mat) +
    theme(
        axis.text.x = element_text(angle = -60, hjust = 0),
        panel.border = element_rect(linetype = "dashed", fill = NA)
    ) +
    anno_right(size = unit(15, "mm")) +
    align_dendro(reorder_dendrogram = TRUE) +
    anno_top(size = unit(15, "mm")) +
    align_dendro(reorder_dendrogram = TRUE) +
    theme(plot.margin = margin())
```

You can use the `filling` argument to turn off the heatmap cell filling,
allowing you to customize the heatmap body geoms. Use the `color` aesthetic to
specify the cell border color and the `linewidth` aesthetic to set the border
width.
```{r}
ggheatmap(mat, filling = NULL) +
    geom_tile(aes(fill = value), color = "white") +
    theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
    anno_right(size = unit(15, "mm")) +
    align_dendro(reorder_dendrogram = TRUE) +
    anno_top(size = unit(15, "mm")) +
    align_dendro(reorder_dendrogram = TRUE) &
    theme(plot.margin = margin())
```

To draw a blank heatmap body:
```{r}
ggheatmap(mat, filling = NULL) +
    theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
    anno_right(size = unit(15, "mm")) +
    align_dendro(reorder_dendrogram = TRUE) +
    anno_top(size = unit(15, "mm")) +
    align_dendro(reorder_dendrogram = TRUE) &
    theme(plot.margin = margin())
```

# Titles
We can use `patch_titles()` to add titles around each border of the plot. You
can use `theme()` to control the text appearance. 
```{r}
ggheatmap(mat) +
    theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
    anno_right(size = unit(20, "mm")) +
    align_dendro(reorder_dendrogram = TRUE) +
    patch_titles(right = "I am a row title") +
    theme(plot.patch_title.right = element_text(face = "bold", size = 16)) +
    anno_top(size = unit(15, "mm")) +
    align_dendro(reorder_dendrogram = TRUE) +
    patch_titles(top = "I am a column title") +
    theme(plot.patch_title.top = element_text(face = "bold", size = 16)) &
    theme(plot.margin = margin())
```

# Clustering
## Distance methods
```{r}
# ComplexHeatmap::Heatmap(mat,
#   name = "mat", clustering_distance_rows = "pearson",
#   column_title = "pre-defined distance method (1 - pearson)"
# )
ggheatmap(mat) +
    theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
    anno_right(size = unit(20, "mm")) +
    align_dendro(distance = "pearson", reorder_dendrogram = TRUE) +
    anno_top(size = unit(15, "mm")) +
    align_dendro(reorder_dendrogram = TRUE) +
    patch_titles(top = "pre-defined distance method (1 - pearson)") +
    theme(plot.patch_title.top = element_text(face = "bold", size = 16)) &
    theme(plot.margin = margin())
```

```{r}
# ComplexHeatmap::Heatmap(mat,
#     name = "mat", clustering_distance_rows = function(m) dist(m),
#     column_title = "a function that calculates distance matrix"
# )
ggheatmap(mat) +
    theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
    anno_right(size = unit(20, "mm")) +
    align_dendro(distance = dist, reorder_dendrogram = TRUE) +
    anno_top(size = unit(15, "mm")) +
    align_dendro(reorder_dendrogram = TRUE) +
    patch_titles(top = "a function that calculates distance matrix") +
    theme(plot.patch_title.top = element_text(face = "bold", size = 16)) &
    theme(plot.margin = margin())
```

## Clustering methods
Method to perform hierarchical clustering can be specified by `method` argument,
Possible methods are those supported in `hclust()` function. 
```{r}
# ComplexHeatmap::Heatmap(mat,
#     name = "mat",
#     clustering_method_rows = "single"
# )
ggheatmap(mat) +
    theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
    anno_right(size = unit(20, "mm")) +
    align_dendro(method = "single", reorder_dendrogram = TRUE) +
    anno_top(size = unit(15, "mm")) +
    align_dendro(reorder_dendrogram = TRUE) &
    theme(plot.margin = margin())
```

Use `distance = NULL` if you don't want to calculate the distance.
```{r eval=rlang::is_installed("cluster")}
ggheatmap(mat) +
    theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
    anno_right(size = unit(20, "mm")) +
    align_dendro(
        distance = NULL, method = cluster::diana,
        reorder_dendrogram = TRUE
    ) +
    anno_top(size = unit(15, "mm")) +
    align_dendro(
        distance = NULL, method = cluster::agnes,
        reorder_dendrogram = TRUE
    ) &
    theme(plot.margin = margin())
```

## Render dendrograms
It's easy for `ggalign` to color the branches by setting the color mapping,
since `ggalign` will add the `cutree()` results into the underlying data.
```{r}
ggheatmap(mat) +
    theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
    anno_right(size = unit(20, "mm")) +
    align_dendro(aes(color = branch), k = 2L, reorder_dendrogram = TRUE) +
    anno_top(size = unit(15, "mm")) +
    align_dendro(reorder_dendrogram = TRUE) &
    theme(plot.margin = margin())
```

# Set row and column orders
We can use `align_order()` to set the order.
```{r}
# ComplexHeatmap::Heatmap(mat,
#     name = "mat",
#     row_order = order(as.numeric(gsub("row", "", rownames(mat)))),
#     column_order = order(as.numeric(gsub("column", "", colnames(mat)))),
#     column_title = "reorder matrix"
# )
ggheatmap(mat) +
    theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
    anno_right(size = unit(20, "mm")) +
    align_order(as.numeric(gsub("row", "", rownames(mat)))) +
    anno_top(size = unit(15, "mm")) +
    align_order(as.numeric(gsub("column", "", colnames(mat)))) &
    theme(plot.margin = margin())
```

# Seriation
`align_reorder()` can directly take the `seriate()` function as the input and
extract the ordering information.
```{r eval=rlang::is_installed("seriation")}
mat2 <- max(mat) - mat
ggheatmap(mat2) +
    scale_fill_gradient2(low = "blue", high = "red", midpoint = 2L) +
    theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
    anno_right(size = unit(20, "mm")) +
    align_reorder(seriation::seriate, method = "BEA_TSP") +
    anno_top(size = unit(15, "mm")) +
    align_reorder(seriation::seriate, method = "BEA_TSP") &
    theme(plot.margin = margin())
```

The above code will execute `seriate()` twice—once for each dimension. However,
since a single run of `seriate()` can provide the ordering for both dimensions,
we can manually extract the ordering indices to avoid redundancy.
```{r eval=rlang::is_installed("seriation")}
o <- seriation::seriate(mat2, method = "BEA_TSP")
ggheatmap(mat2) +
    scale_fill_gradient2(low = "blue", high = "red", midpoint = 2L) +
    theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
    anno_right(size = unit(20, "mm")) +
    align_order(seriation::get_order(o, 1L)) +
    anno_top(size = unit(15, "mm")) +
    align_order(seriation::get_order(o, 2L)) &
    theme(plot.margin = margin())
```
For more use of the `seriate()` function, please refer to the
[seriation](https://cran.r-project.org/web/packages/seriation/index.html)
package. 

# Dimension labels
`ggplot2` use scales and theme to control the axis labels, Please see
[scales](https://yunuuuu.github.io/ggalign/articles/scales-and-facets.html#scales)
section for more details.
```{r}
# ComplexHeatmap::Heatmap(mat,
#     name = "mat", row_names_side = "left", row_dend_side = "right",
#     column_names_side = "top", column_dend_side = "bottom"
# )
ggheatmap(mat) +
    scale_x_continuous(position = "top") +
    scale_y_continuous(position = "right") +
    theme(axis.text.x = element_text(angle = 60, hjust = 0)) +
    anno_left(size = unit(15, "mm")) +
    align_dendro(reorder_dendrogram = TRUE) +
    scale_x_continuous(position = "top") +
    anno_bottom(size = unit(15, "mm")) +
    align_dendro(reorder_dendrogram = TRUE) +
    scale_y_continuous(position = "right") +
    quad_active() &
    theme(plot.margin = margin())
```

```{r}
ggheatmap(mat) +
    scale_y_continuous(breaks = NULL) +
    theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
    anno_right(size = unit(15, "mm")) +
    align_dendro(reorder_dendrogram = TRUE) +
    anno_top(size = unit(15, "mm")) +
    align_dendro(reorder_dendrogram = TRUE) +
    quad_active() &
    theme(plot.margin = margin())
```

```{r}
ggheatmap(mat) +
    theme(
        axis.text.x = element_text(angle = -60, hjust = 0),
        axis.text.y = element_text(face = "bold", size = 16)
    ) +
    anno_right(size = unit(15, "mm")) +
    align_dendro(reorder_dendrogram = TRUE) +
    anno_top(size = unit(15, "mm")) +
    align_dendro(reorder_dendrogram = TRUE) +
    quad_active() &
    theme(plot.margin = margin())
```

```{r}
ggheatmap(mat) +
    theme(
        axis.text.x = element_text(angle = -60, hjust = 0),
        axis.text.y = element_text(
            face = "bold", size = 16,
            colour = c(rep("red", 10), rep("blue", 8))
        )
    ) +
    anno_right(size = unit(15, "mm")) +
    align_dendro(reorder_dendrogram = TRUE) +
    anno_top(size = unit(15, "mm")) +
    align_dendro(reorder_dendrogram = TRUE) +
    quad_active() &
    theme(plot.margin = margin())
```

# Heatmap split
## Split by k-means clustering
```{r}
ggheatmap(mat) +
    theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
    anno_right(size = unit(15, "mm")) +
    align_kmeans(2L) +
    align_dendro(reorder_dendrogram = TRUE) +
    anno_top(size = unit(15, "mm")) +
    align_dendro(reorder_dendrogram = TRUE) +
    quad_active() &
    theme(plot.margin = margin())
```

```{r}
ggheatmap(mat) +
    theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
    anno_right(size = unit(15, "mm")) +
    align_dendro(reorder_dendrogram = TRUE) +
    anno_top(size = unit(15, "mm")) +
    align_kmeans(3L) +
    align_dendro(reorder_dendrogram = TRUE) +
    quad_active() &
    theme(plot.margin = margin())
```

```{r}
ggheatmap(mat) +
    theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
    anno_right(size = unit(15, "mm")) +
    align_kmeans(2L) +
    align_dendro(reorder_dendrogram = TRUE) +
    anno_top(size = unit(15, "mm")) +
    align_kmeans(3L) +
    align_dendro(reorder_dendrogram = TRUE) +
    quad_active() &
    theme(plot.margin = margin())
```

The dendrogram was calculated in each group defined by kmeans.

## Split by categorical variables
```{r}
ggheatmap(mat) +
    theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
    anno_right(size = unit(15, "mm")) +
    align_group(rep(c("A", "B"), 9)) +
    align_dendro(reorder_dendrogram = TRUE) +
    anno_top(size = unit(15, "mm")) +
    align_group(rep(c("C", "D"), 12)) +
    align_dendro(reorder_dendrogram = TRUE) +
    quad_active() &
    theme(plot.margin = margin())
```

## Split by dendrogram
When you splitted by a dendrogram, the cutted height will be indicated with a
dashed line.
```{r}
ggheatmap(mat) +
    theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
    anno_right(size = unit(15, "mm")) +
    align_dendro(k = 3L, reorder_dendrogram = TRUE) +
    anno_top(size = unit(15, "mm")) +
    align_dendro(k = 2L, reorder_dendrogram = TRUE) +
    quad_active() &
    theme(plot.margin = margin())
```

```{r}
ggheatmap(mat) +
    theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
    anno_right(size = unit(15, "mm")) +
    align_dendro(aes(color = branch), k = 3L, reorder_dendrogram = TRUE) +
    scale_color_brewer(palette = "Dark2") +
    anno_top(size = unit(15, "mm")) +
    align_dendro(k = 2L, reorder_dendrogram = TRUE) +
    quad_active() &
    theme(plot.margin = margin())
```

## Order of slices (panels)
The order of the panels always follow the factor level. Note: the merging of
dendrogram between `ComplexHeatmap` and `ggalign` is a little different.
```{r}
ggheatmap(mat) +
    theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
    anno_right(size = unit(15, "mm")) +
    align_group(rep(LETTERS[1:3], 6)) +
    align_dendro(aes(color = branch),
        reorder_dendrogram = TRUE,
        reorder_group = TRUE,
        merge_dendrogram = TRUE
    ) +
    scale_color_brewer(palette = "Dark2") +
    anno_top(size = unit(15, "mm")) +
    align_group(rep(letters[1:6], 4)) +
    align_dendro(aes(color = branch),
        reorder_dendrogram = TRUE,
        reorder_group = TRUE,
        merge_dendrogram = TRUE
    ) +
    quad_active() -
    with_position(theme(strip.text = element_text()), "tr") &
    theme(plot.margin = margin())
```

```{r}
ggheatmap(mat) +
    theme(axis.text.x = element_text(angle = -60, hjust = 0)) +
    anno_right(size = unit(15, "mm")) +
    align_group(rep(LETTERS[1:3], 6)) +
    align_dendro(aes(color = branch), reorder_dendrogram = TRUE) +
    scale_color_brewer(palette = "Dark2") +
    anno_top(size = unit(15, "mm")) +
    align_group(rep(letters[1:6], 4)) +
    align_dendro(aes(color = branch), reorder_dendrogram = TRUE) +
    quad_active() -
    with_position(theme(strip.text = element_text()), "tr") &
    theme(plot.margin = margin())
```

## Titles for splitting (facet strip text)
By default, the facet strip text is removed. You can override this behavior with
`theme(strip.text = element_text())`. Since `align_group()` does not create a
new plot, the panel title can only be added to the heatmap plot.

## Graphic parameters for splitting
```{r}
ggh4x::facet_grid2(strip = ggh4x::strip_themed(
    background_x = list(
        element_rect(fill = "red"),
        element_rect(fill = "blue"),
        element_rect(fill = "green")
    )
))
```

# Session information
```{r}
sessionInfo()
```
